DNA replication errors are a major source of adaptive gene amplification

Version 1

Abstract

Abstract

Detecting and understanding heritable changes in DNA that contribute to adaptive evolution is a primary goal of evolutionary biology. Copy number variants (CNVs)—gains and losses of genomic sequences—are an important source of genetic variation underlying rapid adaptation and genome evolution. However, despite their central role in evolution little is known about the factors that contribute to the structure, size, formation rate, and fitness effects of adaptive CNVs. Local genome elements are likely to be an important determinant of these properties. Whereas it is known that point mutation rates vary with genomic location and local sequence features, the role of genome architecture in the formation, selection, and the resulting evolutionary dynamics of CNVs is poorly understood. Previously, we have found that the GAP1 gene in Saccharomyces cerevisiae undergoes frequent and repeated amplification and selection under long-term experimental evolution in glutamine-limiting conditions. The GAP1 gene has a unique genomic architecture consisting of two flanking long terminal repeats (LTRs) and a proximate origin of DNA replication (autonomously replicating sequence, ARS), which are likely to promote rapid GAP1 CNV formation. To test the role of these genomic elements on CNV-mediated adaptive evolution we performed experimental evolution in glutamine-limited chemostats using engineered strains lacking either the adjacent LTRs, ARS, or all elements. Using a CNV reporter system and neural network simulation-based inference (nnSBI) we quantified the formation rate and fitness effect of CNVs for each strain. We find that although GAP1 CNVs repeatedly form and sweep to high frequency in all strains, removal of local DNA elements significantly impacts the rate and fitness effect of CNVs and the rate of adaptation. We performed genome sequence analysis to define the molecular mechanisms of CNV formation for 177 CNV lineages. We find that 49% of all GAP1 CNVs are mediated by the DNA replication-based mechanism Origin Dependent Inverted Repeat Amplification (ODIRA) regardless of background strain. In the absence of the local ARS, a distal ARS can mediate ODIRA CNV formation. In the absence of local LTRs homologous recombination mechanisms still mediate gene amplification following de novo insertion of retrotransposon elements at the locus. Our study demonstrates the remarkable plasticity of the genome and reveals that DNA replication errors are a predominant source of adaptive CNVs.

Introduction

Defining the genetic basis and evolutionary dynamics of adaptation is a central goal in evolutionary biology. Mutations underlying adaptation or biological innovation can depend on multiple factors including genetic backgrounds, phenotypic states, and genome architecture (Blount et al., 2008, 2012). One important class of mutation mediating adaptive evolution are copy number variants (CNVs) which comprise duplications or deletions of genomic sequences that range in size from gene fragments to whole chromosomes. Quantifying the rates at which CNVs occur, the factors that influence their formation, and the fitness and functional effects of CNVs is essential for understanding their role in evolutionary processes.

CNVs play roles in rapid adaptation in multiple contexts and are an initiating event in biological innovation. In laboratory evolution experiments, a spontaneous tandem duplication captured a promoter for expression of a citrate transporter and resulted in Escherichia coli cells, typically unable to use citrate, to start metabolizing citrate as a carbon source (Blount et al., 2012). CNVs can be beneficial in cancer cells, promote tumorigenesis (Ben-David & Amon, 2020), enhance cancer cell adaptability (Rutledge et al., 2016), and accelerate resistance to anti-cancer therapies (Lukow et al., 2021). Over longer time scales, CNVs serve as substrate from which new genes evolve (Ohno, 1970; Taylor & Raes, 2004) as duplicated genes redundant in function can accumulate mutations and evolve to acquire new functions. For example, the globin gene family in mammals arose from rounds of gene duplication and subsequent diversification (Storz, 2016). CNVs also contribute to macro-evolutionary processes and thereby contribute to species differences, such as between humans and chimpanzees (Cheng et al., 2005) and speciation (Zuellig & Sweigart, 2018).

Mutations, including CNVs, occur in part because of errors made during DNA replication or DNA repair. Two general processes underlie CNV formation: (1) DNA recombination-based mechanisms and (2) DNA replication-based mechanisms (Harel et al., 2015; Hastings, Lupski, et al., 2009; Malhotra & Sebat, 2012; Pös et al., 2021; Zhang, Gu, et al., 2009). Recombination-mediated mechanisms of CNV formation include non-allelic homologous recombination (NAHR) and nonhomologous end joining. NAHR occurs via recombination between homologous sequences that are not allelic. As such, NAHR occurs more frequently with repetitive sequences due to improper alignment of DNA segments and can occur either between (interchromosomal) or within (intrachromosomal) a chromosome (Harel et al., 2015). One prevalent class of repetitive sequence are retrotransposons and both full length and partial sequences, such as long terminal repeats (LTR), are substrates for homologous recombination generating gene amplifications (Avecilla et al., 2023; Dunham et al., 2002; Gresham et al., 2008; Lauer et al., 2018; Spealman et al., 2022). DNA replication-based mechanisms include fork stalling template switching (FoSTes) and microhomology mediated break-induced repair (MMBIR) (Carvalho et al., 2013; Gu et al., 2008a; Hastings, Ira, et al., 2009; Lee et al., 2007). During FoSTes and MMBIR, the DNA replication fork stalls due to a single strand nick and a replication error occurs in which the lagging strand switches to an incorrect template strand mediated by microhomology. Reinitiation of DNA synthesis at the incorrect site can form CNVs. A particular type of DNA replication-based error is Origin-Dependent Inverted Repeat Amplification (ODIRA), in which short inverted repeats near an origin of DNA replication enable template switching of the leading strand to the lagging strand. Subsequent replication generates an intermediate linear DNA molecule that can recombine into the original genome to form a triplication with an inverted middle copy (Brewer et al., 2015; Martin et al., 2024).

In microbes, CNVs can mediate rapid adaptation to selective conditions imposed through nutrient limitation in a chemostat. Selected CNVs often include genes encoding nutrient transporters that facilitate import of the limiting nutrient (Dunham et al., 2002; Gresham et al., 2008; Horiuchi et al., 1963; Payen et al., 2016; Sonti & Roth, 1989), likely as a result of improved nutrient transport capacity due to increased protein production. Previous studies have found amplification of the general amino acid permease gene, GAP1, when Saccharomyces cerevisiae populations are continuously cultured in glutamine-limited chemostats (Gresham et al., 2010; Lauer et al., 2018). Amplification of GAP1 confers increased fitness in the selective environment (Avecilla et al., 2023). Sequence characterization of these CNVs revealed that a diversity of de novo CNV alleles arose and were selected for including tandem duplications of GAP1, complex large CNVs, aneuploidies, and translocations. However, little is known about the molecular basis of this genetic diversity.

Local genome elements are likely to be an important determinant of CNV formation rates and mechanisms. Genomic context can influence multiple properties including mutation rate, epigenetic regulation, chromatin state, transcription levels, DNA replication, and recombination rate (Arndt et al., 2005; Chuang & Li, 2004; Lang & Murray, 2011; Lercher & Hurst, 2002; Matassi et al., 1999; Nishant et al., 2009; Wolfe et al., 1989). Prior work has shown that CNVs occur more frequently in repetitive regions in the genome (Harel et al., 2015; Pentao et al., 1992; Stankiewicz et al., 2003; Turner et al., 2008). However, little is known about the role of local genomic architecture and organization on CNV formation rates, the types of CNVs that are generated, their associated fitness effects, and ultimately the paths taken during adaptive evolution.

Here, we aimed to directly investigate the effect of local genome architecture elements on de novo GAP1 CNV formation and selection dynamics during adaptive evolution of Saccharomyces cerevisiae. We hypothesized that sequence elements proximate to GAP1 potentiate CNV formation. The GAP1 locus, which is located on the short arm of chromosome XI, consists of two flanking Ty1 long terminal repeats (LTRs) that share 82% sequence identity and an origin of DNA replication or autonomously replicating sequence (ARS) (Figure 1A). Both LTRs and ARS may facilitate GAP1 CNV formation due to their proximity. First, the flanking LTRs can undergo inter-chromatid NAHR to form tandem duplications of GAP1 on a linear chromosome (Lauer et al., 2018; Spealman et al., 2022). Second, intra-chromatid NAHR between the flanking LTRs can form an extrachromosomal circle containing GAP1 and an ARS able to self-propagate and integrate into the genome (Gresham et al., 2010). Finally, GAP1 triplications can form through ODIRA using short inverted repeats and the proximate ARS (Brewer et al., 2015; Lauer et al., 2018; Martin et al., 2024). These elements are thought to facilitate a high rate of GAP1 amplification, estimated to be on the order of 10-4 per haploid genome per generation (Avecilla et al., 2022). To test our hypothesis we used a CNV reporter, wherein a constitutively expressed fluorescent GFP gene is inserted adjacent to GAP1 (Lauer et al., 2018). We engineered strains that lacked either the ARS (ARSΔ), both flanking LTRs (LTRΔ), or all three elements (ALLΔ) (Figure 1A). We performed experimental evolution using wildtype (WT) and genomic architecture mutant populations in glutamine-limited chemostats for 137 generations and quantified GAP1 CNVs using flow cytometry (Figure 1). Surprisingly, we find that the proximate DNA elements are not required for GAP1 CNV formation as GAP1 CNVs were identified in all evolving populations. We used neural network simulation-based inference (nnSBI) to infer the CNV formation rate and selection coefficient (Avecilla et al., 2022). We find that genomic architecture mutants have significantly reduced CNV formation rates relative to WT and significantly lower selection coefficients even though GAP1 CNVs repeatedly form and sweep to high frequency in all strains. We performed genome sequence analysis to define the molecular mechanisms of CNV formation for 177 CNV lineages and found that 49% of GAP1 CNVs are mediated by ODIRA regardless of background strain. In the absence of the local ARS, a distal ARS facilitates CNV formation through ODIRA. We also find that homologous recombination mechanisms still mediate gene amplification in the absence of LTRs in part initiated by de novo insertion of retrotransposon elements at the locus. Our study reveals the remarkable plasticity of the genome and that DNA replication errors are a predominant source of adaptive CNVs.

Local ARS contributes to CNV dynamics during adaptive evolution.

(A) The Saccharomyces cerevisiae GAP1 gene is located on the short arm of chromosome XI (beige rectangle). Light blue rectangle - Ty Long terminal repeats (LTR). Purple rectangle - Autonomously replicating sequences (ARS). Orange rectangles - tRNA genes. GAP1 ORF - white rectangle. The GAP1 gene (white rectangle) is flanked by Ty1 LTRs (YKRCδ11, YKRCδ12), which are remnants of retrotransposon events and is directly upstream of an autonomously replicating sequence (ARS1116). Variants of the GAP1 locus were engineered to remove either both LTRs, the single ARS, or all three elements. All engineered genomes contain a CNV reporter. (B) We evolved the four different strains in 5-8 replicate populations, for a total of 27 populations, in glutamine-limited chemostats and monitored the formation and selection of de novo GAP1 CNVs for 137 generations using flow cytometry. Population samples were taken every 8-10 generations and 100,000 cells were assayed using a flow cytometer. Colored lines show the median proportion of cells in a population with GAP1 amplifications across 5-8 replicate populations of the labeled strain. The shaded regions represent the median absolute deviation across the replicates. (C) We summarized CNV dynamics and found that strain has a significant effect on CNV appearance (Kruskal-Wallis, p =0.001384). There are significant differences in CNV appearance between LTRΔ (blue) and ARSΔ (red), and LTRΔ (blue) and ALLΔ (yellow) (pairwise wilcoxon test with Bonferroni correction, p=0.0059 and p=0.0124, respectively). (D) Strain has a significant effect on the per generation increase in proportion of cells with CNV (ANOVA, p = 0.00318). There is a significant difference between LTRΔ (blue) and ALLΔ (yellow) (pairwise t-test with Bonferroni correction, p = 0.0026). (E) Strain has a significant effect on time to CNV equilibrium phase (ANOVA, p = 0.00833). There is a significant difference in time to CNV equilibrium between WT and ARSΔ (pairwise t-tests with bonferroni correction, p =0.050).

Results

Accurate estimation of CNV allele frequencies remains challenging using molecular methods such as DNA sequencing and qPCR. To address this challenge we developed a CNV reporter comprising a constitutively expressed fluorescent gene inserted upstream of GAP1 and observed recurrent amplification and selection of GAP1 in glutamine-limiting conditions (Lauer et al., 2018). Subsequently, we showed that a high rate of GAP1 CNV formation and strong fitness effects explain the highly reproducible evolutionary dynamics (Avecilla et al., 2022). Noncoding sequence elements proximate to GAP1, including flanking LTRs in tandem orientation and an ARS, contribute to GAP1 CNV formation (Gresham et al., 2010; Lauer et al., 2018). Many studies have shown that repetitive sequence regions and origins of replications are hotspots of CNVs (Arlt et al., 2012; Cardoso et al., 2016; Di Rienzi et al., 2009; Gresham et al., 2010; Lauer et al., 2018; Martin et al., 2024). Thus, we hypothesized that the local genomic architecture of GAP1 facilitates its high rate of CNV formation.

To test the role of proximate genomic features we engineered strains deleted for each element and thus differing from our wildtype strain (WT) containing a GAP1 CNV reporter by a single modification. Specifically, we constructed ARSΔ, a strain lacking the single ARS, LTRΔ, a strain lacking the flanking LTRs, and ALLΔ, a strain lacking all three elements (Figure 1A). All strains contain the CNV reporter at the identical location as the WT strain. We confirmed scarless deletions of genetic elements using Sanger and whole-genome sequencing.

Local genomic architecture contributes to GAP1 CNV evolutionary dynamics

We founded independent populations with each of the three engineered strains lacking proximate genomic features and a WT strain. We studied GAP1 CNV dynamics in populations maintained in glutamine-limited chemostats over 137 generations (Figure 1). For each of the four strains, we propagated 5-8 clonal replicate populations, each originating from the same inoculum (founder population) derived from a single colony. Approximately every 10 generations, we measured GFP fluorescence of sampled populations using a flow cytometer and quantified the proportion of cells containing GAP1 CNVs (Methods). We observed similar CNV dynamics across independent populations within each strain (Figure S3B). Therefore, we summarized CNV dynamics for each strain using the median proportion of the population with a GAP1 CNV. In every strain, GAP1 CNVs are generated and selected resulting in qualitatively similar dynamics in WT and mutant strains (Figure 1B).

Deletion of the ARS, but not the flanking LTRs alters CNV dynamics

We quantified three phases of CNV dynamics 1) the time to CNV appearance, defined by the inflection point before rise in frequency (Fig 1C), 2) the selection phase, corresponding to the increase in proportion of CNVs per generation during the initial expansion of CNVs (Figure 1D), and 3) the equilibrium phase, corresponding to the plateau (Figure 1E). The time to CNV appearance (Figure 1C) and the CNV selection phase (Figure 1B) does not differ between WT and LTRΔ populations (pairwise wilcoxon test, adjusted p = 1, pairwise t-test p = 1, respectively). In the WT and LTRΔ populations, GAP1 CNVs appear at generation 50 and increase in frequency at similar rates, ∼15% per generation in WT and ∼18% per generation in LTRΔ. The two strains both reach their equilibrium phase at the same point around generation 75 (pairwise t-test, adjusted p =1). The absence of a significant difference in CNV adaptation dynamics between the two strains suggests that the LTRs are not a major determinant of GAP1 CNV evolutionary dynamics.

By contrast, in ARSΔ and ALLΔ populations, we observe a delay in the time to CNV appearance. In both of these strains, CNVs are first detected at generations 65-80, whereas in WT and LTRΔ populations CNVs are first detected at generation 50 (ARSΔ vs. LTRΔ, wilcoxon pairwise test, adjusted p = 0.0059, ALLΔ vs. LTRΔ, wilcoxon pairwise test, adjusted, p = 0.0124) (Figure 1). Thus, the local ARS contributes to the initial GAP1 CNV dynamics. Similarly, CNV selection is significantly different between the LTRΔ (18%) and ALLΔ (13%) (pairwise t-test, adjusted p-value = 0.0026). Finally, we also observe a significant delay (ANOVA, p = 0.00833) in the generation at which the CNV frequency reaches equilibrium in ARSΔ (∼generation 112) compared to WT (pairwise t-test, adjusted p = 0.05) (Figure 1E). These observations suggest that absence of the ARS in the ARSΔ and ALLΔ strains delays the appearance of GAP1 CNVs compared with the presence of the ARS in WT or LTRΔ strains.

GAP1 amplifications can occur without CNV reporter amplification

In both WT and LTRΔ populations we observed that GAP1 CNV abundance stabilized around 75% during the equilibrium phase (Figure 1B). In these populations, inspection of raw flow cytometry data revealed a persistent single-copy GFP subpopulation (Figure 2A). These data could be explained by two possible scenarios: (1) the existence of a non-GAP1 CNV subpopulation comprising beneficial variation at other loci with fitness effects equivalent to GAP1 CNVs or (2) lineages with GAP1 CNVs without co-amplification of the CNV reporter. To resolve these two possibilities we sequenced clones from the single-copy GFP subpopulation in each of the five WT populations (Supplementary table 1) and identified the presence of GAP1 amplifications without co-amplification of the CNV reporter (Supplementary figure S2A). The presence of GAP1 CNVs in four out of five populations lacking the CNV reporter amplification suggests it may have occurred in the founder population. We found eight distinct CNV breakpoint pairs (Supplementary figure S2A) indicating at least eight independent events occurred, either in the founder population or just after being propagated to independent populations. All clones from one of the five populations (i.e. population 3) contained one copy of GFP and one copy of GAP1, suggesting these clones have a beneficial mutation elsewhere in the genome allowing them to stably coexist with the GAP1 CNV subpopulation.

CNV reporter failure does not impact parameter inference.

(A) Flow cytometry of a representative WT population with a persistent single-copy GFP subpopulation comprising ∼25% of the population. (B) Model illustration. XA is the frequency of ancestor cells in the chemostat; XC+, XC− are the frequencies of cells with GAP1 duplications with two or one reporters, respectively, and a selection coefficient sC; XB is the frequency of cells with other beneficial mutations and a selection coefficient sB. GAP1 duplications form with a rate δC, other beneficial mutations occur with rate δB. At generation 0, only genotypes C− and A are present, with frequencies of XC− = φ and XA = 1 − φ. (C) Examples of total CNV proportions (solid) and reported CNV proportions (dashed) for two parameter combinations, both with sC = 0. 15, φ = 10−4.

Incorporating unreported pre-existing CNVs in an evolutionary model

To quantify the evolutionary parameters underlying empirically measured CNV dynamics (Figure 1) we built an evolutionary model with the goal of performing network simulation-based inference (nnSBI) (Avecilla et al., 2022; Cranmer et al., 2020; Gonçalves et al., 2020). Previously, our evolutionary model assumed the GAP1 CNV reporter allowed us to detect all GAP1 CNVs (Avecilla et al., 2022). However, our new experimental results indicate the existence of a small subpopulation of unreported GAP1 CNVs present at the beginning of the experiments (Figure 2A). Therefore, we expanded the evolutionary model to include φ, the proportion of cells with GAP1 CNVs without co-amplification of the reporter, at the commencement of the experiment (i.e. generation 0) (Figure 2B). The remaining model parameters are δC, the rate at which GAP1 duplications form, and δB, the rate other beneficial mutations occur. We find that this updated evolutionary model can accurately describe the observed dynamics (Supplementary Figure S3B), which are clearly affected by the value of φ. When the total CNV proportion is very different from the reported proportion, e.g., when φ≫δC > δB, a reduced CNV formation rate results in a greater discrepancy between reported and total CNV proportions (Figure 2C)

Decreased CNV formation rates in mutants suggest adjacent elements can drive GAP1 CNV formation

We used nnSBI to estimate CNV formation rates and selection coefficients from evolutionary dynamics observed in glutamine-limited chemostats. We trained a neural density estimator using evolutionary simulations (Methods). This neural density estimator then allows us to infer posterior distributions over model parameters (formation rate, δ; selection coefficient, s) from the CNV dynamics in single replicates and a collective posterior distribution from a set of replicates of the same strain. δC is the GAP1 CNV formation rate. sC is the GAP1 CNV selection coefficient, wherein the fitness effect is 1 + s. We estimated the confidence of our inference approach on synthetic simulations by computing its coverage, i.e., the probability that the true parameter falls within the 95% highest density interval (HDI) of the posterior distribution. Our neural density estimator is slightly over-confident for φ with a coverage of 0.934, and under-confident for GAP1 CNV selection coefficient and formation rate with a coverage of 0.992 for sC and 0.995 for δC, respectively. (Supplementary Table 2). Despite this under-confidence, the posterior distributions are narrow in biological terms: the 95% HDI represents less than an order of magnitude for both sC and δC. Thus, we did not apply post-training adjustments to the neural density estimator, such as calibration (Cook et al., 2006) or ensembles (Caspi et al., 2023; Hermans et al., 2022).

We find that the individual maximum a posteriori (MAP) estimates vary across strains and replicates (Supplementary Figure S3A). Overall, the CNV selection coefficient, sC, ranges from 0.1 to 0.22 (with one exception of 0.3) whereas the CNV formation rate, δC, ranges from 10−6 to 10−4 (with one exception of 10−3 and two of 10−7); and the proportion of pre-existing GAP1 CNVs that do not amplify the reporter φ ranges from 10−6 to 10−2 (with two exceptions of 10−8). We found that the MAP estimates for replicate populations within the same strain cluster together with some outliers (Supplementary Figure S3A). We performed posterior predictive checks, drawing parameter values from the posterior distributions and simulating the frequency dynamics (Supplementary Figure S3H), which agree with the observed data (Supplementary Figure S3B). For each strain, we estimate the collective posterior distribution based on all individual posteriors, which generates a posterior distribution conditioned on all observations, P(θ|X1, …, Xn) (Methods). The collective posterior allows us to estimate whether there is a difference in CNV formation rate and fitness effects across the four strains.

Collective posterior HDIs are very narrow (Figure 3A) and fit empirical observations (Figure 3C). The collective MAP estimates of the CNV selection coefficient are similar for the WT and LTRΔ (0. 182). For ARSΔ and ALLΔ, the selection coefficient is estimated to be lower, with values of 0.146 and 0.126, respectively. However, these selection coefficients are still large, consistent with these populations containing GAP1 CNVs that are highly beneficial under glutamine-limitation. The collective MAP estimate for the CNV formation rate in WT is 4. 5⋅10−5. In contrast with selection coefficients, the CNV formation rate is markedly lower in all strains ranging from 1⋅10−5 for LTRΔ and ALLΔ to 2. 4⋅10−6 in ARSΔ. These results are consistent with our hypothesis that proximate sequence features facilitate GAP1 CNV formation.

Inference of CNV formation rate and selection coefficient from experimental evolutionary data.

(A) Collective MAP estimate (black markers) and 50% HDR (colored areas) of GAP1 CNV formation rate, δC, and selection coefficient, sC. Marginal posterior distributions are shown on the top and right axes. (B) Collective posterior prediction of Shannon diversity of CNV lineages (e−Σi[pilog(pi)], Jost, 2006). Line and shaded area show mean and 50% HDI. (C) CNV reported frequency (XC+) prediction using collective MAP (solid line) compared to empirical observations (dotted lines).

In all strains, collective parameter estimations are highly correlated, as expected for joint estimation of selection coefficients and beneficial mutation rates (Gitschlag et al., 2023). Thus, the collective MAP predictions can be interpreted as a range of parameter values for each strain. Indeed, other than the very final time point for the WT population, all collective MAP predictions lay within the interquartile ranges (Supplementary Figure S3B). The observed GAP1 CNV frequency stabilizes at different levels in the different experiments (Figure 1B). This can be explained by pre-existing unreported CNVs with frequencies estimated to be between φ=4⋅10−6 to 1. 6⋅10−4 by the collective MAPs (Supplementary Figure S3C). Indeed, our model predicts that the total (reported and unreported) final CNV frequency is nearly one in all cases (Supplementary Figure S3G).

Next, we estimated the de novo CNV diversity. Previous work showed a diversity of CNV allele types formed under glutamine-limited selection including tandem duplications, segmental amplification, translocations, and whole chromosome amplification (Lauer et al., 2018), and that lineage richness decreases rapidly over the course of evolution due in part to competition and clonal interference (Lauer et al., 2018; Levy et al., 2015; Nguyen Ba et al., 2019). Our model does not include competition, clonal interference, or recurrent CNV formation. Therefore, diversity calculations are likely overestimations. Nonetheless, a comparison of diversity between strains is informative of whether proximate genome elements affect CNV allele diversity. Therefore, for each strain, we used its collective MAP to simulate a posterior prediction for the genotype frequencies (Figure 3C), which we then used to predict the posterior Shannon diversity (Jost, 2006). We predict the set of CNV alleles to be highly diverse: the final predicted Shannon diversity ranges from 1. 6⋅104 in ARSΔ to 3. 2⋅105 in WT (Figure 3B). Our model predicts that the diversity increases rapidly during the selection phase and stabilizes in the equilibrium phase. This is because CNV alleles that form towards the end of the experiment would have a low frequency with a minor effect on diversity. We observe the greatest diversity in WT populations with lower diversity in the three genomic architecture mutants. Moreover, diversity saturates faster in WT populations. This suggests that the WT strain is able to form more unique CNVs allele types earlier compared to the other three strains (Figure 3B). Shannon diversity is lower in LTRΔ and further lower in ALLΔ and ARSΔ (Figure 3B). This rank order is the same as the CNV formation rates (Figure 3A).

We used a modified version of the evolutionary model with the inferred parameters to simulate an evolutionary competition between WT and the three architecture mutant strains over 116 generations, a point in which CNVs have reached high frequency in the original experiment. To “win” these competitions, the competitor strains need to adapt to glutamine-limitation by producing CNVs (Supplementary Figure S3J and supplementary information). The results of the simulated evolutionary competitions predict that the WT would dominate over all three strains, as its predicted final proportion almost always exceeds its initial frequency of 0.5. The average predicted proportion of WT cells when competing with LTRΔ is 0.717. In contrast, ARSΔ and ALLΔ are predicted to be almost eliminated by generation 116, as the average predicted WT proportion is 0.998 and 0.999, respectively. These simulated competitions further suggest that the ARS is a more important contributor to adaptive evolution mediated by GAP1 CNVs.

Inference of CNV mechanisms in genome architecture mutants

Contrary to our expectations, removal of proximate genomic elements from the GAP1 locus does not inhibit the formation of GAP1 CNVs. We sought to determine the molecular basis by which GAP1 CNVs form in the absence of these local elements. Therefore, we isolated ∼40 GAP1 CNV-containing clones from each population containing the four different strains at generations 79 and 125 and performed Illumina whole-genome sequencing. Using a combination of read depth, split read, and discordant read analysis, we defined the extent of the amplified region, the precise CNV breakpoints, and GAP1 copy number. On the basis of these features, we inferred the CNV-forming mechanisms for each GAP1 CNV (Methods). Over the 177 analyzed GAP1 CNVs, we observed tandem amplifications, tandem triplications with an inverted middle copy, intra-chromosomal translocations, aneuploidy, and complex CNVs. GAP1 copy numbers range from two to six in any given clone. Each of the four strains is able to produce a diversity of CNV alleles ranging from small (tens of kilobases) to large (∼hundreds of kilobases) segmental amplifications (Figure 4). We identified six major CNV-forming mechanisms across the four strains: ODIRA, LTR NAHR, NAHR, transposon-mediated, complex CNVs, and whole chromosome duplication (aneuploidy) (Figure 4A and Methods).

Local and distal elements contribute to generation of GAP1 CNV alleles.

(A) Schematic of Saccharomyces cerevisiae GAP1 locus on Chromosome XI: 513332-518060 with LTR, ARS elements and tRNA genes labeled. ODIRA is a DNA replication-error based CNV mechanism. Here, we classify a clone as ODIRA if it has an inverted sequence in at least one breakpoint. ODIRA typically forms tandem triplications with an inverted middle copy and contains an ARS. Long terminal repeat non-allelic homologous recombination (LTR NAHR) is a mechanism we define by having both CNV breakpoints at LTR sites. Sometimes we detect a hybrid sequence between two LTR sequences, a result of recombination between the two LTRs. Non-allelic homologous recombination (NAHR) is defined by having at least one CNV breakpoint not at LTR sites, ie. other homologous sequences in the genome. Sometimes we detect a hybrid sequence between the two homologous sequences. Transposon-mediated mechanisms observed involve at least one intermediate novel LTR retrotransposon insertion. The newly deposited LTR sequences recombines with other LTR sequences, either pre-existing or introduced by a second de novo retrotransposition, to form a resulting CNV. Complex CNV is defined by a clone having more than two breakpoints in chromosome XI, indicative of having more than one amplification event. (B) Violin plot of CNV length in each genome-sequenced clone, n = 177. Strain has a significant effect on CNV length, Kruskal-Wallis test, p = 3.0 x 10-4. (C) Barplot of inferred CNV mechanisms, described in A, for each CNV clone isolated from evolving populations. Inference came from a combination of read depth, split read, and discordant read analysis (see Methods). Strain is significantly associated with CNV Mechanism Fisher’s Exact Test, p = 5.0 x 10-4. There is a significant increase in ODIRA prevalence between WT and LTRΔ, chi-sq, p = 0.02469. There is a significant decrease in ODIRA prevalence from WT to ARSΔ and ALLΔ, chi-sq, p = 0.002861 and 0.002196, respectively. There is a significant decrease of LTR NAHR from WT to LTRΔ, chi-sq, p = 0.03083.

(D) Top: Schematic of S. cerevisiae chromosome XI, with LTR, ARS elements, tRNA genes annotated. LTR-blue, ARS-purple, tRNA-orange, GAP1 ORF-white rectangle. Using a combination of read depth, split read, and discordant read analysis, we defined the extent of the amplified region, the precise CNV breakpoints, and GAP1 copy number. GAP1 copy numbers were estimated using read depth relative to the average read depth of chromosome XI. We define the upstream and downstream breakpoints as kilobases away from the start codon of the GAP1 ORF (vertical dotted line). Bottom: Dumbbell plots represent the amplified region (>1 copy) relative to the WT architecture reference genome. The ends of the dumbbells mark the approximate CNV breakpoints. Select clones were chosen as representative of the observed diversity of amplifications.

(E) Scatterplots of CNV length for all genome-sequenced clones, n = 177. We defined the upstream and downstream breakpoints as kilobases away from the start codon of the GAP1 ORF (vertical dotted line in dumbbell plot). CNV mechanisms are defined in Figure 4A.

ODIRA is a predominant mechanism of CNV formation

We inferred GAP1 CNVs formed through ODIRA in all four genotypes at high frequencies: 22 out of 37 WT clones (59%), 42 out of 52 LTRΔ clones (81%), 11 out of 42 ARSΔ clones (26%), and 12 out of 46 ALLΔ clones (26%). Considering the set of all CNVs in all strains, ODIRA is the most common CNV mechanism comprising almost half of all CNVs (87/177, 49%). The second most common mechanism is NAHR between flanking LTRs (38/177, 21%), which generate tandem amplifications. In the WT background, ODIRA (22/37) and NAHR between LTRs (11/37) account for 89% of GAP1 CNVs.

In LTRΔ, GAP1 CNVs form via ODIRA, chromosome missegregation, and NAHR using other sites. As expected, in LTRΔ clones we did not detect NAHR between LTRs in 52 clones and as a result focal amplifications were not detected (Figure 4C). In LTRΔ, CNVs are formed predominantly by ODIRA (42/52, 81%) (Supplementary Table 3), a significant increase relative to WT clones (chi-sq, p = 0.02469) (Figure 4C). By contrast, aneuploidy (5/52), complex CNV (3/52), and NAHR (2/52) account for less than 10% of GAP1 CNVs in LTRΔ. Consequently, we observe an increase in GAP1 CNV size in LTRΔ relative to WT (Figure 4B) as there is an increased prevalence of segmental amplifications and aneuploidy (Figure 4E).

Aneuploidy was rarely observed. Whole amplification of chromosome XI was detected in six out of 177 clones (3.4%) (Figure 4C). It was detected only in 2 strains: WT and LTRΔ (Figure 4D).

ODIRA generates CNVs using distal ARS

Whereas removal of proximate LTRs abrogates the formation of CNVs through NAHR, removal of the local ARS does not prevent the formation of GAP1 CNVs through ODIRA (Figure 4C). In the absence of the proximate ARS, distal ones are used to form ODIRA as all amplified regions of ODIRA clones contain a distal ARS (Figure 4D), with one exception (see Methods, Supplementary Figure S4B). We observe an increase of LTR NAHR in the ARSΔ clones (27/52, 52%) relative to WT clones (11/37, 39%) (Figure 4C, chi-sq, p = 0.03083). In ARSΔ, we find two CNV size groups (Figure 4B). Small amplifications are formed via NAHR of the flanking LTRs (Figure 4E) and large segmental amplifications are formed via ODIRA and by NAHR between one local and one distal LTR (Figure 4E).

Novel retrotransposition events potentiate GAP1 CNVs

CNVs in the ALLΔ clones form by two major mechanisms: 1) ODIRA using distal ARS sites to form large amplifications and 2) LTR NAHR following novel Ty LTR retrotransposon insertions to form focal amplifications (transposon-mediated, Figure 4). ALLΔ clones have larger amplifications formed by ODIRA than ODIRA-generated amplification in WT and LTRΔ (Figure 4E) because they encompass distal ARS and inverted repeats (Figure 4D). Surprisingly, we detected novel LTR retrotransposon events that generated new LTRs that subsequently formed GAP1 CNVs through NAHR with a pre-existing LTR in the genome or an LTR from a second novel retrotransposition (Figure 4E). Regions upstream of tRNA genes are known to be hotspots for Ty retrotransposons (Ji et al., 1993; Mularoni et al., 2012). We find the novel retrotransposons insert near one or both of the previously deleted LTR sites (Supplementary File 1), which flank GAP1 and are downstream of tRNA genes (Figure 1A). We only detected novel retrotranspositions in ALLΔ populations. In total we detected 15 unique Ty retrotransposon insertion sites of which eight were upstream of the deleted LTR, YKRCδ11, and four were downstream of deleted LTR, YKRCδ12 (Supplementary File 1). The remaining two insertions were distal to the GAP1 gene: one on the short arm and the second on the long arm of chromosome XI. Every insertion was upstream of an tRNA gene, consistent with the biased preference of Ty LTR insertions (Ji et al., 1993; Mularoni et al., 2012). Recombination between a new and preexisting LTR produces large amplifications whereas recombination between two newly inserted Ty1 flanking the GAP1 gene forms focal amplifications of the GAP1 gene (Figure 4E).

Discussion

In this study we sought to understand the molecular basis of repeated de novo amplifications and selection of the general amino acid permease gene, GAP1, in S. cerevisiae continuously cultured in glutamine-limited selection. We hypothesized that a high formation rate of GAP1 CNVs is due to the unique genomic architecture at the locus, which comprises two flanking long terminal repeats and a DNA replication origin. We used genetic engineering, experimental evolution, and neural network simulation-based inference to quantify de novo CNV dynamics and estimate the CNV formation rate and selection coefficient in engineered mutants lacking the proximate genome elements. We find that removal of these elements has a significant impact on de novo CNV dynamics, CNV formation rate, and selection coefficients. However, CNVs are formed and selected in the absence of these elements highlighting the plasticity of the genome and diversity of mechanisms that generate CNVs during adaptive evolution.

Despite their proximity to GAP1 we found that flanking LTRs are not an essential driver of CNV formation. The de novo CNV dynamics of WT and LTRΔ populations are similar and we find that although the CNV formation rate is reduced, the effect is small. In contrast, a significantly decreased CNV formation rate and delayed CNV appearance time was observed in the absence of the ARS in ARSΔ and ALLΔ populations, which suggests that the local ARS is a major determinant of GAP1 CNV-mediated adaptive dynamics. Indeed, ODIRA was identified as the predominant CNV mechanism in sequence-characterized clones revealing that DNA replication errors are a major source of CNV formation during adaptive evolution.

The prevalence of ODIRA is a result of many replication origins and the pervasiveness of inverted repeat sequences throughout the chromosome (Figure 4). In particular, breakpoint analysis of LTRΔ CNV clones show that ODIRA produces a continuum of CNV sizes along the short arm of chromosome XI. Downstream breakpoints range from near the GAP1 gene (∼3 kilobases) to all the way to the right telomere of chromosome XI (153 kilobases) (Figure 4E). The S. cerevisiae genome contains a high number, 1-1000 per kilobase, of inverted repeats ranging from 3bp to 14bp (Martin et al., 2024). The longer repeats are more likely to be used in ODIRA (Martin et al., 2024). The ubiquity of inverted repeats is in stark contrast to the relative paucity of LTR sequences, which are dispersed throughout the genome. Thus, ODIRA supplies a diverse and high number of gene amplifications for selection to act on, setting the stage for genome evolution and adaptation.

Consistent with previous reports of increased Ty insertions in S. cerevisiae under stress conditions (Morillon et al., 2000, 2002), we observed novel retrotransposon insertions in populations evolved in glutamine-limited chemostats. Transposon insertions could be harmful and lead to loss-of-function mutations but are also substrate for generating beneficial alleles including CNVs (Blanc & Adams, 2003; Dunham et al., 2002; Gresham et al., 2008; Wilke & Adams, 1992). We only detected novel Ty insertions in the ALLΔ strain. This is likely because regions upstream of tRNA genes unoccupied by LTRs are predisposed to transposition. Our detection of novel retrotransposon insertions is consistent with a previous experimental evolution study that suggested that Ty insertions were rare under constant nitrogen-limitation and substantially more common under fluctuating nitrogen limitation, in which cells experience total nitrogen starvation periodically (Hays et al., 2023). We observe a substantially lower frequency of novel Ty retrotransposition events, 15/177 clones, compared to 898/345 clones, with multiple insertions per genome, isolated from serial transfer ammonium-limited experimental evolution (Hays et al., 2023). Importantly, the role of Ty differs in the two studies, as in our case beneficial CNV formed after novel retrotransposition through recombination of newly introduced LTR sequences, whereas Hays et al. found Ty associated null alleles that are beneficial in nitrogen-limited conditions. Our results reveal the prevalence of Ty insertions even under constant nitrogen-limiting conditions and demonstrate how new retrotransposition events can contribute to CNV formation.

Aneuploidy was not a major source of adaptation in our experiments as it was infrequently detected (n = 6/177). This contrasts with studies suggesting aneuploidy is a rapid and transient route to adaptation over short evolutionary time scales (Chen, Bradford, et al., 2012; Chen et al., 2015; Chen, Rubinstein, et al., 2012; Pavelka et al., 2010; Selmecki et al., 2015; Yona et al., 2012).

However, aneuploidy incurs a fitness cost (Robinson et al., 2023; Tsai et al., 2019; Yang et al., 2021) and therefore can be outcompeted by slow-forming but less costly beneficial mutations in large populations (Kohanovski et al., 2024). Our observations of higher frequencies of focal and segmental amplifications may be because they are less costly than whole-chromosome amplifications.

A variety of DNA replication errors generate CNVs. Replication slippage at palindromic DNA and DNA repeats can cause fork stalling and downstream CNV formation (Lee et al., 2007; Zhang, Khajavi, et al., 2009). DNA repeats can form secondary structures like R loops, cruciforms, non-B DNA structures, and hairpins which stimulate CNV formation (Gu et al., 2008b). Untimely replication, faulty fork progression, S-phase checkpoint dysfunction, defective nucleosome assembly, and DNA repeat sites including LTRs are sources of replication-associated genome instability (Aguilera & García-Muse, 2013). However, DNA replication alone might not be the only contributor. The GAP1 gene is highly transcribed under glutamine-limitation (Airoldi et al., 2016). Transcription-replication collisions may fuel ODIRA CNV formation at this locus (Lauer & Gresham, 2019; Wilson et al., 2015). CNV formation can also be stimulated by transcription-associated replication stress and histone acetylation (Hull et al., 2017; Salim et al., 2021; Whale et al., 2022) and replication fork stalling at tRNA genes (Osmundson et al., 2017; Yeung & Smith, 2020). Testing the role of transcription in promoting the formation of adaptive CNVs warrants further investigation.

Recent work has proposed that ODIRA CNVs are a major mechanism of CNVs in human genomes (Brewer et al., 2015; Martin et al., 2024). Studies of human and yeast genomes have typically considered homologous recombination as the predominant mechanism of CNV formation (Lupski & Stankiewicz, 2005). CNV hotspots identified in the human (Chance et al., 1994; Lupski, 1998; Lupski & Stankiewicz, 2005; Pentao et al., 1992) and yeast genomes are indeed mediated by NAHR of long repeat sequences (Green et al., 2010; Gresham et al., 2010). However, a focus on recombination-based mechanisms as a means of generating copy number variation may be the result of ascertainment bias or the comparative ease of studying the effect of long repeat sequences over short palindromic ones. Our study demonstrates that experimental evolution in yeast is a useful approach to elucidating the molecular mechanisms by which DNA replication errors generate CNVs.

Methods

Strains and Media

Each of the three architecture mutants were constructed independently starting with the GAP1 CNV reporter strain (DGY1657). To construct each deletion strain, we performed two rounds of transformations both using PCR amplified donor templates designed for homology-directed repair. The first transformation used a repair template containing a nourseothricin resistance cassette to replace the pre-existing kanamycin resistance cassette and GAP1 gene. The repair template was designed to also deleted the elements of interests (ie. ARS, both flanking LTRs, or both LTRs and ARS). The second transformation replaced the nourseothricin cassette with a kanamycin resistance cassette and GAP1 gene thus yielding a genomic architecture Δ strain. We confirmed scarless deletions with sanger sequencing and whole-genome-sequencing. Final identifiers are DGY1657 for the WT architecture strain, DGY2076 for the LTRΔ strain, DGY2150 for the ARSΔ strain, and DGY2071 for the ALLΔ strain.

DGY1, DGY500, and DGY1315 are zero-, one-, two-, copy GFP controls, respectively, described in Lauer et al, 2018 and Spealman et al. 2023 (Lauer et al., 2018; Spealman et al., 2023). Briefly, GFP under the ACT1 promoter was inserted at neutral loci that do not undergo amplification in glutamine-limited continuous culture. 400μM glutamine-limited media is described in Lauer et al. 2018.

Long-Term Experimental Evolution

We performed experimental evolution of 30 S. cerevisiae populations in miniature chemostats (ministats) for ∼137 generations under nitrogen limitation with 400μM glutamine as in Lauer et al. (2018). Of the 30 populations, there were three controls: one control population with no fluorescent reporter (DGY1), one with one GFP fluorescent reporter (DGY500), one with two GFP fluorescent reporters (DGY1315). The remaining 27 populations have the GAP1 CNV GFP reporter. Of these, five populations are WT (DGY1657), seven are LTRΔ (DGY2076), seven are ARSΔ (DGY2150), and eight are ALLΔ (DGY2071). We inoculated each ministat containing 20ml of glutamine-limited media with 0.5 ml culture from its corresponding genotype founder population. The founder population was founded by a single colony grown overnight in glutamine-limited media at 30°C. Replicate populations of the same strain were inoculated from the same founder population derived from a single colony. Strains were randomized among the 30-plex ministat setup to account for the possibility of systematic position effects. After inoculation, populations were incubated in a growth chamber at 30°C for 24 hours with the media inflow pump off. After 24 hours, the populations had reached early stationary phase and we turned on the media inflow pump and waited 4 hours for the populations to reach steady-state equilibrium, at which the population size was ∼108 cells. This was generation zero. Ministats were incubated in a growth chamber at 30°C with a dilution rate of 0.12 culture volumes/hr. Since the ministats had a 20ml culture volume, the population doubling time was 5.8 hours. Approximately every 10 generations, we froze 2-ml samples of each population in 15% glycerol stored at -80°C. Approximately every 30 generations, we pelleted cells from 1-ml samples of each population and froze them at -80°C for genomic DNA extraction.

Flow Cytometry analysis to study GAP1 CNV dynamics

To track GAP1 CNV dynamics, we sampled 1-ml from each population approximately every 10 generations. We sonicated cell populations for 1 minute to remove any cell clumping and immediately analyzed samples on the Cytek Aurora flow cytometer. We sampled 100,000 cells per population and recorded forward scatter, side scatter, and GFP fluorescent signals for every cell. We performed hierarchical gating to define cells, single cells, unstained (zero-copy-GFP control) cells, cells with one copy of GFP (GAP1), and two or more copies of GFP (GAP1) (Spealman et al., 2023). First we gated for cells (filtered out any debris, bacteria) by graphing forward scatter area (FSC-A) against side scatter area (SSC-A). Second, we gated for single cells by graphing forward scatter area against forward scatter height and drawing along the resulting diagonal. Finally, we drew non-overlapping gates to define three subpopulations: zero copy, one copy, and two or more copies of GFP by graphing B2 channel area (B2-A), which detects GFP (excitation = 516 λ, emission = 529 λ), against forward scatter area (FSC-A). We note that the one copy and two copy events overlap some, which is a limitation in this experiment (Spealman et al., 2023).

We found that two architecture mutants, DGY2150 and DGY2071, had strain-specific GFP fluorescence even though they only harbored one copy of GFP. DGY2150 and DGY2071 had slightly higher fluorescence than the one copy GFP control strain, DGY500, but less than that of the two copy GFP control strain, DGY1315. The third architecture mutant, DGY2076, had the same GFP fluorescence as the one-copy GFP control strain (DGY500). We ruled out that they were spontaneous diploids by looking at forward scatter signals. The forward scatter signal was not different from that of the one copy control (a haploid) and was not as high as a diploid. Therefore due to strain-specific fluorescence, we decided to perform strain-based gating, ie. one set of gates for the WT strain, a second set of gates for the LTRΔ strain, and so on. Since the controls are also a strain of their own, they were not used to set universal gates for one-copy or two-copy. Thus, for each strain, we chose the basis of our one-copy gate as the timepoint per strain in aggregate with the lowest median cell-sized normalized fluorescence. The two-or-more-copy (CNV) gate was drawn directly above and non-overlapping with the one-copy gate.

Quantification of dynamics

To obtain the proportion of CNVs for each population at each timepoint, we applied gates that correspond to zero-,one-, and two-or-more copy subpopulations. Using such proportion per population per timepoint, we summarize population CNV dynamics as follows (Lauer et al., 2018; Spealman et al., 2023). We calculate the generation of CNV appearance for each of the evolved populations. We defined CNV appearance as the generation where the proportion of CNV-containing cells first surpasses a threshold of 10% for three consecutive generations. Next, modified from Lang et al. (2011) (Lang et al., 2011) and Lauer et al. (2018) (Lauer et al., 2018), we calculate the percent increase in CNVs per generation for each evolved population. We compute the natural log of the proportion of the population with CNVs divided by the proportion of the population without CNVs for each timepoint. These proportions were obtained previously by gating. We plot these values across time and perform linear regression during the initial increase of CNVs. The slope of the linear regression is the percent increase in CNVs per generation. Finally, we calculate the time to CNV equilibrium, as defined by the generation at which a linear regression results in a slope < 0.005 after the selection phase.

Neural network simulation-based inference of evolutionary parameters

Evolutionary model

We developed a Wright-Fisher model that describes the evolutionary dynamics, similar to our previous study (Avecilla et al., 2022). In this study we have shown that a Wright-Fisher model is suitable for describing evolutionary dynamics in a chemostat. This is a discrete-time, non-overlapping generations model with a constant population size. Every generation has three stages: selection, in which the frequency of genotypes with beneficial alleles increases; mutation, in which genotypes can gain a single beneficial mutation or CNV; and drift, in which the population of the next generation is generated by sampling from a multinomial distribution. Our model follows the change in frequency of four genotypes (Figure 2B): A, the ancestor genotype; B, a cell with a non-CNV beneficial mutation; C+, a genotype with two copies of GAP1 and two copies of the CNV reporter; and C−, a genotype with two copies of GAP1 but only a single copy of the CNV reporter. CNV and non-CNV alleles are formed at a rate of δC and δB and have a selection coefficient of sC and sB, respectively. The frequency of genotype i is Xi. Unlike XB and XC+, which may increase due to both mutation and selection, we assume that C− is not generated after generation 0 (as experimental results suggest that the reporter is working properly). Hence, the C− genotype only increases in frequency due to selection, with sC as its selection coefficient. We assume C− has an initial frequency φ. Model equations and further details are in the supplementary information.

Simulation-based inference

We use a neural network simulation-based inference method, Neural Posterior Estimation or NPE (Papamakarios, 2019) to estimate the joint posterior distribution of three model parameters, sC, δC and φ, while the other parameters, sB and δB are fixed to a specific value(Table 1). Inferring all five model parameters resulted in similar prediction accuracy and sC and δC estimates.

We applied NPE implemented in the Python package sbi (Tejero-Cantero et al., 2020) using a masked autoregressive flow (Papamakarios et al., 2018) as the neural density estimator: an artificial neural network that “learns” an amortized posterior of model parameters from a set of synthetic simulations. Posterior amortization allows us to estimate the posterior distribution P(θ|Χ) for a new observation Χ without the need to re-run the entire inference pipeline, i.e., generating new simulations and re-training the network (as is the case in sampling-based methods such as Markov chain Monte Carlo or MCMC).

We generated 100,000 synthetic observations simulated from our evolutionary model using parameters drawn from the prior distribution (Table 1). The neural density estimator was trained using early stopping with a convergence threshold of 100 epochs without decreases in minimal validation loss (the default in sbi is 20). Using 100 epochs as a threshold resulted in improved predictions. We validated that this improvement in prediction accuracy is not a result of over-fitting (Supplementary Figure S3F).

We validated the trained neural density estimator by measuring the coverage property: the probability of parameters to fall within the estimated posterior marginal 95% HDI. Then, we used the distribution of (Supplementary Figure S3D) and posterior predictive checks (Supplementary Figure S3B) as quantitative and qualitative measures of prediction accuracy, respectively.

Collective posterior distribution

NPE estimates a posterior distribution per observation, i.e., P(θ|X). Given n observations X1, …Xn generated from the same model distribution P(θ), where each observation is a time-series of GAP1 CNV frequency, NPE infers n individual posterior distributions, each conditioned on a single observation, P(Xi). We estimate the collective posterior distribution based on n individual posteriors, that is, a posterior distribution conditioned on all observations, This can be computed using the individual posteriors P(Xi) and the prior P(θ) (see supplementary information for derivation).

However, as P(Xi) could be infinitesimally small, a single observation could potentially reject a parameter value that is likely according to other observations. We want the collective posterior to be robust to such non-representative observations. Therefore, we define P∈(Xi) = max(∈, P(Xi)) and use this quantity instead of P(Xi) in eq. 1. For a correct choice of ∈, the collective posterior mode should reflect a value with high posterior density for multiple observations, rather than a value that no individual posterior completely rejects. Thus, we implemented an amortized collective posterior: given a set of observations, X1, …Xn, an amortized posterior for individual observations, P(θ|X), a prior, P(θ), and ∈, a collective posterior distribution P(θ|X1, …, Xn), can be computed. This distribution can then be used to evaluate the posterior distribution of specific parameters given the observations, generate random samples from the posterior distribution (e.g., using rejection sampling), or find the collective MAP.

We set ∈ = e−150 based on a visual grid-search. To find the normalizing factor (denominator in eq. 1), the integral is approximated by a dense Riemann sum (3003 points). Maximizing the distribution, i.e., finding the collective MAP, is implemented using scipy’s minimize method with the Nelder-Mead algorithm.

Genetic diversity

Using our evolutionary model with the inferred parameters, we can estimate the diversity of CNV alleles in the experiments. For each strain, we used samples from its collective posterior to simulate a posterior prediction for the CNV allele frequencies (Figure 3C), which we then used to compute the posterior Shannon diversity (Jost, 2006), as detailed in the supplementary information.

Whole genome sequencing of isolated clones

Clones were isolated from archived populations and verified to harbor a GAP1 CNV by measuring GFP fluorescence signal consistent with two or more copies. Populations of each strain from generation 79 were streaked out from the -80°C archive on YPD and incubated at 30°C for 2 days. Plates containing single colonies were viewed under a blue light to view GFP fluorescent colonies by eye. Relative to the fluorescence of the 2 copy control strain, we picked single colonies that fluoresced as bright or brighter, reasoning that these colonies would likely contain GAP1 CNVs. Single colonies were used to inoculate cultures in glutamine-limited media and incubated at 30°C for 18 hours. The cultures were analyzed on the Cytek Aurora to verify they indeed harbored two or more copies of GAP1 based on GFP fluorescence signal. For Illumina whole genome sequencing, genomic DNA was isolated using Hoffman-Winston method. Libraries were prepared using a Nextera kit and Illumina adapters. Libraries were sequenced on Illumina NextSeq 500 platform PE150 (2 x 150 300 Cycle v2.5) or Illumina NovaSeq 6000 SP PE150 (2x 150 300 Cycle v1.5). We also used custom Nextera Index Primers reported in S1 Table Baym et al. 2015 (Baym et al., 2015).

Breakpoint analysis and CNV mechanism inference in sequenced clones

Reference genomes

We created a custom reference genome for each of the genomic architecture mutants. The custom reference genome containing the GAP1 CNV reporter in Lauer et al. 2018 (NCBI assembly R64) was modified to delete the flanking LTRs, single ARS, or all three elements.

Copy number estimation by read depth

The estimation of GAP1 copy number from read depth used is described in Lauer et al. 2018, except we searched for ≥ 1000 base pairs of contiguous sequence. CNV boundaries were refined by visual inspection.

Structural variation calling and breakpoint analysis

Whole genome sequences of clones were run through CVish, a structural variant caller (Spealman, 2019). Structural variant calling was also done on each of the ancestor genomes: WT, ARSΔ, LTRΔ, and ALLΔ. Output .bam files containing split reads and discordant reads of evolved clones and its corresponding ancestor were visualized on Jbrowse 2 to confirm locations of de novo CNV breakpoints and orientation of sequences at the novel breakpoint junctions. Novel contigs relative to the reference genome were outputted in addition to the supporting split reads that generated the contig. Blastn was used to verify orientation of contigs, namely inverted sequences used to define ODIRA (see Definitions of Inferred CNV Mechanisms).

Definitions of Inferred CNV mechanisms

We used the following liberal classifications for each CNV category. We called a clone ODIRA if we found inverted sequences in at least one breakpoint (Figure 4A). We define LTR NAHR as having both breakpoints at LTR sites (Figure 4A). This is evidence of recombination between the homologous LTR sequences. This mechanism typically forms tandem amplifications. In some cases, we find the hybrid sequence between two LTRs, but this is hard to retrieve in short-read-sequencing.

We define NAHR as having breakpoints at homologous sequences, with at least one breakpoint not at an LTR sequence (Figure 4A). We define transposon-mediated as a clone having a breakpoint at a novel LTR retrotransposon insertion and the other breakpoint at a different LTR site (Figure 4A). This supports that the newly deposited LTR sequence recombined with other LTR sequences (either pre-existing or introduced by a second de novo retrotransposition) to form CNVs. Sometimes we are able to recover the hybrid sequence between LTR sequences. We define complex CNV as having more than two breakpoints on chromosome XI and a read depth profile that looks like more than one amplification event occurred (multi-step profile). For the complex CNV clones, we were not able to resolve the CNV mechanisms due to the limitations of short-read sequencing.

Supplement

Independent GAP1 amplifications lacking CNV reporter amplification. Read depth plots of the GAP1 CNV reporter locus of sequenced clones from five 1-copy-GFP subpopulations. Identification of eight distinct CNV breakpoint pairs, shown above, across the populations indicate the occurrence of at least eight independent amplifications of GAP1 without GFP amplification. GFP reference gene - green rectangle, GAP1 reference gene - purple rectangle.

MAP estimates of GAP1 CNV formation rates (δC) and selection coefficients (sC) for all replicate populations.

Markers show MAP estimates from individual replicates, crosses show 50% HDI of collective posteriors. Extreme points are marked for comparison to data and posterior prediction, see Supplementary Figure S3B for posterior predictive checks.

Posterior predictive checks for all replicates.

Black markers are the empirical observations, dashed line shows MAP prediction. The leftmost plot of each row shows the collective MAP prediction with empirical data’s interquartile range (gray bars).

Pairwise and marginal collective posteriors for all estimated model parameters.

Diagonals show marginal collective posteriors per parameter per strain. Below-diagonal plots show pairwise KDEs for all pairs of model parameters. Collective joint MAPs (which may differ from collective marginal MAPs, as the marginal distribution integrates over all other parameters), are marked by a red vertical line. Panels are separated by strain: (A) WT, (B) ARSΔ, (C) LTRΔ, (D) ALLΔ.

Parameter estimation accuracy on synthetic data.

Log-ratio of MAP estimate and true parameter value for 829 synthetic simulations in which the final reported GAP1 CNV proportion is at least 0.3.

Neural density estimator training and validation loss during training.

Convergence threshold of 100 unimproved epochs (no decrease in minimal validation loss) was reached after 569 epochs.

Total GAP1 CNV frequency.

Solid lines show collective MAP predictions, dashed lines show the total proportion of GAP1 CNVs, comprising unreported pre-existing CNVs and reported CNVs generated during the experiment, as predicted by the evolutionary model.

Error estimation of parameter inference.

Average root mean square errors (RMSE) of 50 posterior samples against the observed data. (A) Individual posteriors and individual replicates. (B) Collective posterior and individual replicates. (C) Collective posterior and empirical mean.

Pairwise evolutionary competition predictions.

We simulated evolutionary competitions in the experimental conditions of WT vs. genomic architecture mutants, starting from equal frequencies. The proportion at generation 116 of WT was predicted using 10,000 combinations of collective posterior samples for each pairwise competition. Overall, WT outcompetes all mutants because it adapts faster (due to faster CNV formation rate), but its advantage over ARSΔ and ALLΔ is much higher than its advantage over LTRΔ. (A) Histograms for three pairwise competitions. Note that ARSΔ and ALLΔ values overlap at this scale and are all in the rightmost bar. (B) High-resolution histograms for ARSΔ and ALLΔ.

No significant interaction between strain and generation on CNV length.

Boxplot of CNV length of clones by strain and generation of isolation. There is no significant interaction between strain and generation of isolated clone, and no significant effect of generation on CNV length (Two-way ANOVA, Strain x Generation, p = 0.33)

Types of ODIRA detected. We found 87 ODIRA clones total regardless of strain. The majority of ODIRA clones fit the canonical definition of having two inverted junctions and 3 copies, 55/87 clones (63%) (ODIRA_3). We found four non-canonical types. We found 17 clones (20%) with only one inverted junction detected and 3 copies (ODIRA_oneEnd_3). We found 11 clones (13%) with two inverted junctions but only 2 copies (ODIRA_2). We found 3 clones (3.4%) with only one inverted junction detected and 2 copies (ODIRA_oneEnd_2). We found 1 clone (1.1%) with two inverted junctions but the amplified region did not contain an ARS.

Supplementary File 1. Ty-associated clones and locations of novel Ty insertions.

Supplementary File 2. CNV Clone Sequencing Analysis

Data availability

Sequencing data is available at SRA PRJNA1098800.

Other associated data are available here: https://osf.io/js7z8/.

Source code repository simulation-based inference: https://github.com/yoavram-lab/chuong_et_al.

Scripts for flow cytometry-based evolutionary dynamics and analysis of CNV clones: https://github.com/GreshamLab/local_arch_variants.

eLife Assessment

This study provides important new insights into the contribution of local DNA features to the molecular mechanisms and dynamics of copy number variation (CNV) formation during adaptive evolution. While limited to a single CNV, the experiments are carefully controlled and present convincing evidence that supports the conclusions. This work will be of general interest to those studying genome architecture and evolution from yeast biologists to cancer researchers.

Reviewer #1 (Public Review):

Summary:

The work by Chuong et al. provides important new insights into the contribution of different molecular mechanisms in the dynamics of CNV formation. It will be of interest to anyone curious about genome architecture and evolution from yeast biologists to cancer researchers studying genome rearrangements.

Strengths:

Their results are especially striking in that the "simplest" mechanism of GAP1 amplification-non-allelic homologous recombination between the flanking Ty-LTR elements is not the most common route taken by the cells, emphasizing the importance of experimentally testing what might seem on the surface to be obvious answers. One of the important developments of their work is the use of their neural network simulation-based inference (nnSBI) model to derive rates of amplicon formation and their fitness effects.

Weaknesses:

The manuscript reads as though two different people wrote two different sections of the manuscript - an experimental evolutionist and a computational scientist. If the goal is to reach both groups of readers, there needs to be more explanation of both types of work. I found the computational sections to be particularly dense but even the experimental sections need clearer explanations and more specific examples of the rearrangements found. I will point out these areas in the detailed remarks to the authors. While I have no reason to question their conclusions, I couldn't independently verify the results that ODIRA was the majority mechanism since the sequence of amplified clones was not made available during the review. I've encouraged the authors to include specific, detailed sequence information for both ODIRA events as well as the specific clones where GAP1 was amplified but the flanking gene GFP was not.

Reviewer #2 (Public Review):

Summary:

This study examines how local DNA features around the amino acid permease gene GAP1 influence adaptation to glutamine-limited conditions through changes in GAP1 Copy Number Variation (CNV). The study is well motivated by the observation of numerous CNVs documented in many organisms, but difficulty in distinguishing the mechanisms by which they are formed, and whether or how local genomic elements influence their formation. The main finding is convincing and is that a nearby Autonomous Replicating Sequence (ARS) influences the formation of GAP1 CNVs and this is consistent with a predominate mechanism of Origin Dependent Inverted Repeat Amplification (ODIRA). These results along with finding and characterizing other mechanisms of GAP1 CNV formation will be of general interest to those studying CNVs in natural systems, experimental evolution, and in tumor evolution. While the results are limited to a single CNV of interest (GAP1), the carefully controlled experimental design and quantification of CNV formation will provide a useful guide to studying other CNVs and CNVs in other organisms.

Strengths:

The study was designed to examine the effects of two flanking genomic features next to GAP1 on CNV formation and adaptation during experimental evolution. This was accomplished by removing two Long Terminal Repeats (LTRs), removing a downstream ARS, and removing both LTRs and the ARS. Although there was some heterogeneity among replicates, later shown to include the size and breakpoints of the CNV and the presence of an unmarked CNV, both marker-assisted tracking of CNV formation and modeling of CNV rate and fitness effects showed that deletion of the ARS caused a clear difference compared to the control and the LTR deletion.

The consequence of deletion of local features (LTR and ARS) was quantified by genome sequencing of adaptive clones to identify the CNV size, copy number and infer the mechanism of CNV formation. This greatly added value to the study as it showed that i) ODIRA was the most common mechanism but ODIRA is enhanced by a local ARS, ii) non-allelic homologous recombination (NAHR) is also used but depends on LTRs, and iii) de novo insertion of transposable elements mediate NAHR in strains with both ARS and LTR deletions. Together, these results show how local features influence the mechanism of CNV formation, but also how alternative mechanisms can substitute when primary ones are unavailable.

Weaknesses:

The CNV mutation rate and its effect on fitness are hard to disentangle. The frequency of the amplified GFP provides information about mutation rate differences as well as fitness differences. The data and analysis show that each evolved population has multiple GAP1 CNV lineages within it, with some being unmarked by GFP. Thus, estimates of CNV fitness are more of a composite view of all CNV amplifications increasing in frequency during adaptation. Another unknown but potential complication is whether the local (ARS, LTR) deletions influence GAP1 expression and thus the fitness gain of GAP1 CNVs. The neural network simulation-based inference does a good job at estimating both mutation rates and fitness effects, while also accounting for unmarked CNVs. However, the model does not account for the population heterogeneity of CNVs and their fitness effects. Despite these limitations of distinguishing mutation rate and fitness differences, the authors' conclusions are well supported in that the LTR and ARS deletions have a clear impact on the CNV-mediated evolutionary outcome and the mechanism of CNV formation.

Reviewer #3 (Public Review):

Summary:

The authors represent an elegant and detailed investigation into the role of cis-elements, and therefore the underlying mechanisms, in gene dosage increase. Their most significant finding is that in their system copy number increase frequently occurs by what they call replication errors that result from the origin of replication firing.

The authors somewhat quantitatively determine the effect of the presence of a proximal origin of replication or LTR on the different CNV scenarios.

Strengths:

(1) A clever and elegant experimental design.

(2) A quantitative determination of the effect of a proximal origin of replication or LTR on the different CNV scenarios. Measuring directly the contribution of two competing elements.

(3) ODIRA can occur by firing of a distal ARS element.

(4) Re-insertion of Ty elements is interesting.

Weaknesses:

(1) Overall, the research does not considerably advance the current knowledge. The research does not investigate what the maximum distance between ARS for ODIRA is to occur. This is an important point since ODIRA was previously described. A considerable contribution to the field would be to understand under what conditions ODIRA wins NAHR.

(2) The title and some sentences in the abstract give a wrong impression of the generality and the novelty of the observations presented. Below are some examples of much earlier work that dealt with mechanisms of CNV and got different conclusions. The Lobachev lab (Cell 2006) published a different scenario years ago, with a very different mechanism (hair-pin capped breaks). The Argueso lab found something different (NAHR) (Genetics 2013).

In fact, the CUP1 system presents a good example of this point. The Houseley group showed a complex replication transcription-based mechanism (NAR 2022, cited), the Argueso group showed Ty-based amplification and the Resnick group showed aneuploidy-based amplification. While aneuploidy is a minor factor here the numerous works in Candida albicans, Cryptococcus neoformans, and Yeast suggest otherwise (Selmecki et al Science 2006, Yona et al PNAS 2013, Yang et al Microbiology Spectrum 2021).

(3) The authors added a mathematical model to their experimental data. For me, it was very difficult to understand the contribution of the model to the research. I anticipated, for example, that the model would make predictions that would be tested experimentally. For example, " ARSΔ and ALLΔ are predicted to be almost eliminated by generation 116, as the average predicted WT proportion is 0.998 and 0.999" But to my understanding without testing the model.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The work by Chuong et al. provides important new insights into the contribution of different molecular mechanisms in the dynamics of CNV formation. It will be of interest to anyone curious about genome architecture and evolution from yeast biologists to cancer researchers studying genome rearrangements.

Thank you for recognizing the broad significance of our study.

Strengths:

Their results are especially striking in that the "simplest" mechanism of GAP1 amplification-non-allelic homologous recombination between the flanking Ty-LTR elements is not the most common route taken by the cells, emphasizing the importance of experimentally testing what might seem on the surface to be obvious answers. One of the important developments of their work is the use of their neural network simulation-based inference (nnSBI) model to derive rates of amplicon formation and their fitness effects.

We agree with this assessment as the results of our study challenge our intuition that the simplest path to structural variation is the most likely and reveals the great diversity in mechanisms that can lead to large scale changes in the genome.

Weaknesses:

The manuscript reads as though two different people wrote two different sections of the manuscript - an experimental evolutionist and a computational scientist. If the goal is to reach both groups of readers, there needs to be more explanation of both types of work. I found the computational sections to be particularly dense but even the experimental sections need clearer explanations and more specific examples of the rearrangements found. I will point out these areas in the detailed remarks to the authors. While I have no reason to question their conclusions, I couldn't independently verify the results that ODIRA was the majority mechanism since the sequence of amplified clones was not made available during the review. I've encouraged the authors to include specific, detailed sequence information for both ODIRA events as well as the specific clones where GAP1 was amplified but the flanking gene GFP was not.

We have revised the manuscript to expand explanations of both the experimental and computational aspects of our study and to provide additional information for the reader. In doing so, we have edited the text to improve readability. We have made all raw data publicly available through the NCBI short read archive (SRA) and are hosting all sequence data for easy visualization in JBrowse using a public server.

Reviewer #2 (Public Review):

Summary:

This study examines how local DNA features around the amino acid permease gene GAP1 influence adaptation to glutamine-limited conditions through changes in GAP1 Copy Number Variation (CNV). The study is well motivated by the observation of numerous CNVs documented in many organisms, but difficulty in distinguishing the mechanisms by which they are formed, and whether or how local genomic elements influence their formation. The main finding is convincing and is that a nearby Autonomous Replicating Sequence (ARS) influences the formation of GAP1 CNVs and this is consistent with a predominate mechanism of Origin Dependent Inverted Repeat Amplification (ODIRA). These results along with finding and characterizing other mechanisms of GAP1 CNV formation will be of general interest to those studying CNVs in natural systems, experimental evolution, and in tumor evolution. While the results are limited to a single CNV of interest (GAP1), the carefully controlled experimental design and quantification of CNV formation will provide a useful guide to studying other CNVs and CNVs in other organisms.

Thank you for this positive assessment of our study.

Strengths:

The study was designed to examine the effects of two flanking genomic features next to GAP1 on CNV formation and adaptation during experimental evolution. This was accomplished by removing two Long Terminal Repeats (LTRs), removing a downstream ARS, and removing both LTRs and the ARS. Although there was some heterogeneity among replicates, later shown to include the size and breakpoints of the CNV and the presence of an unmarked CNV, both marker-assisted tracking of CNV formation and modeling of CNV rate and fitness effects showed that deletion of the ARS caused a clear difference compared to the control and the LTR deletion.

The consequence of deletion of local features (LTR and ARS) was quantified by genome sequencing of adaptive clones to identify the CNV size, copy number and infer the mechanism of CNV formation. This greatly added value to the study as it showed that i) ODIRA was the most common mechanism but ODIRA is enhanced by a local ARS, ii) non-allelic homologous recombination (NAHR) is also used but depends on LTRs, and iii) de novo insertion of transposable elements mediate NAHR in strains with both ARS and LTR deletions. Together, these results show how local features influence the mechanism of CNV formation, but also how alternative mechanisms can substitute when primary ones are unavailable.

We agree with this assessment.

Weaknesses:

The CNV mutation rate and its effect on fitness are hard to disentangle. The frequency of the amplified GFP provides information about mutation rate differences as well as fitness differences. The data and analysis show that each evolved population has multiple GAP1 CNV lineages within it, with some being unmarked by GFP. Thus, estimates of CNV fitness are more of a composite view of all CNV amplifications increasing in frequency during adaptation. Another unknown but potential complication is whether the local (ARS, LTR) deletions influence GAP1 expression and thus the fitness gain of GAP1 CNVs. The neural network simulation-based inference does a good job at estimating both mutation rates and fitness effects, while also accounting for unmarked CNVs. However, the model does not account for the population heterogeneity of CNVs and their fitness effects. Despite these limitations of distinguishing mutation rate and fitness differences, the authors' conclusions are well supported in that the LTR and ARS deletions have a clear impact on the CNV-mediated evolutionary outcome and the mechanism of CNV formation.

While it is true that the inferred mutation rate and fitness effect are negatively correlated, as in other studies (Gitschlag et al., 2023; Caspi et al., 2023; Avecilla et al., 2022), our modeling approach does generate an estimate of each parameter that is best explained by the data. By reporting the confidence intervals (i.e. the 95% HDI) we define the set of parameter values that are consistent with the data. It is true that our model doesn't explicitly account for population heterogeneity; rather, following Hegreness et al. (2006), we employ a single effective fitness effect and mutation rate for all GAP1 CNVs. It is interesting to consider whether the ARS and LTR affect GAP1 expression; however, we have no evidence that this is the case.

Reviewer #3 (Public Review):

Summary:

The authors represent an elegant and detailed investigation into the role of cis-elements, and therefore the underlying mechanisms, in gene dosage increase. Their most significant finding is that in their system copy number increase frequently occurs by what they call replication errors that result from the origin of replication firing.

The authors somewhat quantitatively determine the effect of the presence of a proximal origin of replication or LTR on the different CNV scenarios.

Strengths:

(1) A clever and elegant experimental design.

(2) A quantitative determination of the effect of a proximal origin of replication or LTR on the different CNV scenarios. Measuring directly the contribution of two competing elements.

(3) ODIRA can occur by firing of a distal ARS element.

(4) Re-insertion of Ty elements is interesting.

We agree that these are interesting and novel findings from our study.

Weaknesses:

(1) Overall, the research does not considerably advance the current knowledge. The research does not investigate what the maximum distance between ARS for ODIRA is to occur. This is an important point since ODIRA was previously described. A considerable contribution to the field would be to understand under what conditions ODIRA wins NAHR.

We agree that these are important questions and they are ones that we are pursuing in future studies.

(2) The title and some sentences in the abstract give a wrong impression of the generality and the novelty of the observations presented. Below are some examples of much earlier work that dealt with mechanisms of CNV and got different conclusions. The Lobachev lab (Cell 2006) published a different scenario years ago, with a very different mechanism (hair-pin capped breaks). The Argueso lab found something different (NAHR) (Genetics 2013).

In fact, the CUP1 system presents a good example of this point. The Houseley group showed a complex replication transcription-based mechanism (NAR 2022, cited), the Argueso group showed Ty-based amplification and the Resnick group showed aneuploidy-based amplification. While aneuploidy is a minor factor here the numerous works in Candida albicans, Cryptococcus neoformans, and Yeast suggest otherwise (Selmecki et al Science 2006, Yona et al PNAS 2013, Yang et al Microbiology Spectrum 2021).

As the reviewer points out there have been several important published studies investigating mechanisms by which structural variation is generated. It is important to note that we are explicitly looking at CNVs in the context of adaptive evolution and the role of genomic features that enable different mechanisms of CNV formation. To emphasize this point, we have changed the title of our manuscript to “Template switching during DNA replication is a prevalent source of adaptive gene amplification”. Aneuploidy is indeed a mechanism of adaptive gene amplification in our current and previously reported studies. We have expanded our discussion to place our study in the context of previous studies reporting mechanisms of gene amplification.

(3) The authors added a mathematical model to their experimental data. For me, it was very difficult to understand the contribution of the model to the research. I anticipated, for example, that the model would make predictions that would be tested experimentally. For example, " ARSΔ and ALLΔ are predicted to be almost eliminated by generation 116, as the average predicted WT proportion is 0.998 and 0.999" But to my understanding without testing the model.

In our previous publication (Avecilla et al. 2022, PLoS Biology) we experimentally validated the use of nnSBI to infer evolutionary parameters. In this study, we have extended our modeling framework to quantify differences between genotypes, which was not previously possible. Our results reveal that the local ARS has a key role in the overall supply rate of CNVs at this locus.

Recommendations for the authors:

We have addressed all public reviews and recommendations.

Reviewer #1 (Recommendations For The Authors):

Specific comments about the work are covered in the order of appearance in the text or Figures. I apologize in advance for the number of comments. They are made out of curiosity, enthusiasm for the research, and a desire to help highlight the most interesting aspects of this work.

We are grateful for the thoughtful comments that have helped us to significantly improve our manuscript.

(1) I would appreciate the inclusion of several references to the work on the ODIRA model.

a) Page 3 last paragraph: "(2) DNA replication-based mechanisms (Harel et al., 2015; Hastings, Lupski, et al., 2009; Malhotra & Sebat, 2012; Pös et al., 2021; Zhang, Gu, et al., 2009; Brewer et al., 2011)" (Addition of Brewer et al., 2011).

We have added all suggested references.

b) Page 4 top: (Brewer et al., 2011; Brewer et al., 2015; Martin et al., 2024). (Addition of Brewer et al., 2011).

We have added all suggested references.

c) Page 14 top: "Recent work has proposed that ODIRA CNVs are a major mechanism of CNVs in human genomes (Brewer et al., 2015; Martin et al., 2024; Brewer et al., 2024)." Brewer et al., 2024 focuses specifically on ODIRA and human CNVs. (Addition of Brewer et al., 2024).

We have added all suggested references.

(2) Page 6, third paragraph: I was surprised that a single inoculating strain was used to establish the replicate chemostats because of the possibility of non-independence of the resulting GAP1 CNVs. A nnSBI model was used to correct for this possibility later in the paper. It seems like it could have been avoided by a simple change in protocol to inoculate each chemostat with an independent inoculum. Was there a reason that the replicate chemostats were not conducted as independent events? Establishing the presence of 'founder' GAP1 CNVs without GFP seems rather secondary to the point of the paper (examining the CNVs that arise during evolution) and I would recommend it being moved to the supplement.

As is typical in microbial experimental evolution studies, we aimed to start with genetically identical homogenous populations and observe the emergence and selection of de novo variation. Therefore, we founded independent populations from a single inoculum. However, this study, and our prior work using lineage tracking barcodes, has clearly demonstrated that during the initial growth of the culture used for the inoculum CNVs are generated that contribute to the adaptation dynamics on all derived populations. This unanticipated result now suggests that the reviewer’s suggestion is a valid one - independent populations should be derived from independent inocula and this will be our standard practice in future studies.

We believe that our results, presented in Figure 2, establishing the presence of pre-existing GAP1 CNVs without the GFP are important as it highlights a limitation of the use of CNV reporters of gene copy number that was not previously known. However, we subsequently show that this class of variant - CNVs that are not detected by the reporter system - can be incorporated into our modeling framework enabling estimation of evolutionary parameters, which we believe is an important finding warranting inclusion in the main text.

(3) Page 7 first full paragraph: "Finally, we also observe a significant delay (ANOVA, p = 0.00833) in the generation at which the CNV frequency reaches equilibrium in ARS∆ (~generation 112) compared to WT (pairwise t-test, adjusted p = 0.05) . . .". Is the delay in reaching a plateau in Figure 1E just a consequence of the later appearance of CNVs or do the authors believe there are two separate events responsible for this delay? E.g. if the authors think that the delay in reaching a plateau is related to lower selection coefficients of the CNVs that do arise compared to the CNVs of other strains, then this should be explicitly discussed.

We believe that the delay in reaching equilibrium is a consequence of both a lower CNV formation and reduced selection coefficients. Lower values for the fitness coefficient and formation rate in ARS∆ explain both the delay in CNV appearance and CNV equilibrium as shown by the predicted dynamics (Figure S3B). We have added an explicit discussion of the effect of the ARS on CNV dynamics in paragraph 2 of the Discussion section paragraph 2 starting at line 456.

(4) Page 7: Incorporating pre-existing CNVs into an evolutionary model: The rationale for how you are able to discount the formation rate of GFP-free CNVs (C-) in your model isn't clear to me. How are you able to assume that these C- events don't form after timepoint 0? Why do you assume a starting population of C- events but not a starting population of C+ events?

We explored the possibility of modeling C- (amplifications of GAP1 without amplification of the reporter) during the evolution experiment. However, because the rate at which C- events occurs is slower than the rate at which C+ events occur (GAP1 amplifications with amplification of the reporter) we found that the effect was negligible. Importantly, the simple model is sufficient to describe the observed dynamics and thus we do not include these possible rare events.

(5) Figure 1:

(a) Panel B: Please put the tRNAs on the line diagrams of the four strains. I first interpreted ALLΔ as missing the tRNAs, too.

Thank you for this suggestion. We added tRNAs to all diagrams to provide additional detail about the structure of the GAP1 locus.

(b) Panels C, D, and E: the dark shade of the colored boxplots obscures the individual points. I recommend reducing the opacity of the box or choosing a lighter shade so that the individual points are visible on top of the box. Is the percent increase in CNVs per generation (Panel D) based on the slopes of the curves in panel B? By eye the slopes of ARS∆ and ALL∆ appear at least as steep as those of wild type and LTR∆.

Thank you for this suggestion. We have now made the individual points visible on top of the boxplots in Figures 1C, 1D, and 1E. The lines in Figure 1B show the median value across populations per time point whereas each point in Figure 1D is the slope from linear regression using values from individual populations (data from individual populations are shown in Figure 3C).

(6) Figure 2:

(a) Panel A: Please remind the readers what FSC-A is measuring and label the different groups of cells in each sample. Are we supposed to assume the upper scatter in generation 8 is the pre-existing CNV variants? Are the three species at generation 50 due to 1, 2, and 3 copies of GFP? Is the new species in generation 137 further amplification of the locus? And if so, how many copies does it represent? I find it fascinating that what I assume is the 2-copy CNV (presumably a direct oriented amplicon produced by NAHR) at 50 generations is lost (out-competed by a potential inverted triplication) at later times, but I didn't find any mention of this phenomenon in the text. What do the different mutant strains look like over the same time course? Please supply supplemental figures with the flow cytometry gating and vertically aligned histograms of the GFP signal so that the peaks are more easily compared. And provide this information for each of the altered strains in supplementary materials.

Thank you for these useful suggestions. We have added a gating legend to the figure to clearly indicate the copy-number for each subpopulation. We have edited the caption and main text to explain forward scatter (FSC-A). Raw flow cytometry plots are now provided as Supplementary figure 2 and distributions of cell-size normalized GFP signal are provided in Supplementary figure 3. Although our primary objective with Figure 2A was to show the persistence of the 1-copy GFP population the reviewer is correct that we did not highlight interesting aspects of the CNV dynamics. We have added additional text starting at line 251 to point out these features of the data.

(b) Panel B: It would help to label the different colored boxes inside cells in Figure 2B - it took me a while to identify the white box as an unrelated adaptive mutation elsewhere in the genome. The linear arrangement of these small colored blocks seems to indicate their structural arrangement. Is that the case? And are they inverted or direct amplicons? Perhaps the authors are being agnostic at this point but it would be better if each of the blocks were separate. If there are other mutations that can explain these GFP-non-amplified survivors, were they identified in your whole genome sequencing?

We have now included a complete legend for Figure 2B indicating that the white box reflects other beneficial mutations. We have separated this class of beneficial mutation from the GAP1 and reporter elements to reflect that they are not linked. We did not identify additional beneficial mutations but plan to pursue this question in a future project.

(c) Panel C: Are the two sets of lines mislabeled? One would expect the "reported" CNV proportions to be lower than the total CNV proportions, not the other way around. Maybe the labels "total CNVs" and "reported CNVs" are unclear to me and I am misunderstanding what "reported" refers to. Please clarify.

Thank you for identifying this mistake. The lines were mislabeled and have now been corrected in the revised version.

(7) Figure 3:

(a) A fuller discussion of panels A and B is needed. The results of panel A in particular seem like an excellent opportunity for connecting the computation to the biology. Can the authors speculate on why the ALL∆ strain has a higher CNV formation rate (𝛿c) than the ARS∆ strain? I would think that taking away one means of amplification would decrease CNV formation. Likewise, could the authors discuss why the selection coefficient (sc) for the LTR∆ strain would be the same as for the wild type? Overall, I would like to see more discussion about what these differences in formation rates and selection coefficients could mean for the types of amplicons arising in the chemostats. (In panel B I don't see the shaded area referred to in the figure legend.) A side-by-side comparison of the data in Panel A with the data shown in Supplemental Figure S3A would be instructive..

Thank you for raising these points. We have added substantial text to the manuscript to address these findings. Starting at line 456 we state:

“The lower CNV formation rate in the LTR∆ could be a closer approximation of ODIRA formation rates at this locus as ODIRA CNVs are the predominant CNV mechanism in the LTR∆ strain (Figure 4F). Furthermore, the low formation rates in the LTR∆ relative to WT might suggest that the presence of the flanking long terminal repeats may increase the rate of ODIRA formation through an otherwise unknown combinatorial effect of DNA replication across these flanking LTRs and template switching at the GAP1 locus. ARS∆ has the lowest CNV formation rate and it could be an approximation of the rates of NAHR between flanking LTRs and ODIRA at distal origins. We find that the ALL∆ has a higher CNV formation rate than the ARS∆, even though three elements are deleted instead of one. One explanation for this is that the deletion of the flanking LTRs in ALL∆ gives opportunity for novel transposon insertions and subsequent LTR NAHR. Indeed we find an enrichment of novel transposon-insertions in the ALL∆ (Figure 4F) and subsequent CNV formation through recombination of the Ty1-associated repeats (Figure 4H, ALL∆). Both events, transposon insertion followed by LTR NAHR, would have to occur quickly at a rate that explains our estimated CNV rate in ALL∆. While remarkable, increased transposon activity has been associated with nutrient stress (Curcio & Garfinkel, 1999; Lesage & Todeschini, 2005; Todeschini et al., 2005) and therefore feasible explanation for the CNV rate estimated in the ALL∆. Additionally, ARS∆ clones rely more on LTR NAHR to form CNVs (Figure 4F). The prevalence of ODIRA in ARS∆ and ALL∆ are similar. LTR NAHR usually occurs after double strand breaks at the long terminal repeats to give rise to CNVs (Argueso et al., 2008). Because we use haploid cells, such double strand break and homology-mediated repair would have to occur during S-phase after DNA replication with a sister chromatid repair template to form tandem duplications. Therefore the dependency on LTR NAHR to form CNVs and the spatial (breaks at LTR sequences) and temporal (S-phase) constraints could explain the lower formation rate in ARS∆.”

In addition, we added a discussion of the different selection coefficients estimated and how the simulated competitions help us understand the decreased selection coefficients in the architecture mutants. In newly added text starting at line 479 we state:

“The genomic elements have clear effects on the evolutionary dynamics in simulated competitive fitness experiments. The similar selection coefficients in WT and LTR∆ suggest that CNV clones formed in these background strains are similar. Indeed, the predominant CNV mechanism in both is ODIRA followed by LTR NAHR (Figure 4F). While LTR NAHR is abolished in the LTR∆, it seems that CNVs formed by ODIRA allow adaptation to glutamine-limitation similar to WT. The lower selection coefficients in ARS∆ and ALL∆ suggest that GAP1 CNVs formed in these strains have some cost. In a competition, they would get outcompeted by CNV alleles in the WT and LTR∆ background.”

(b) The data shown in panel C seems redundant to what is shown more clearly in Supplemental Figure S3B. It seems to me the more important comparison to make in panel C would be the overlay of the predicted data to the median proportion of cells obtained from the experimental data (Figure 1B). Also, overlays of the cultures from each strain could be added to S3A. It is difficult to see the variation within each strain when the data from all four strains are superimposed as they are in Figure 3C.

We agree and have edited Figure 3C to incorporate these suggestions and more clearly convey the intra- and interstrain variation.

(8) Figure 4:

(a) Panels A, B, and C are nice summaries and certainly helpful for understanding panel E, but it would be instructive to see some actual rearrangements of the ODIRA events, the NAHR, and the transposon-mediated rearrangements. It isn't clear to me what these last events look like. A figure that shows the specific architecture of example clones for each category would be helpful. I am also having a hard time reconciling ODIRA events with a copy number of 2. Are these rearrangements free isochromosomes with amplification to the telomere or are they secondary rearrangements like those described in Brewer et al., 2024? And what about the non-aneuploid rearrangement that includes the centromere? Is it a dicentric?

We have now added more detailed depictions of CNVs in Figure 4A and provide links to visualize the alignment files. We have added additional discussion starting at line 397 of the non-canonical ODIRA events and putative neochromosome amplicons with reference to Brewer et al 2024. Starting at line 397 we state:

“Surprisingly, we found CNVs with breakpoints consistent with ODIRA that contained only 2 copies of the amplified region, whereas ODIRA typically generates a triplication. In the absence of additional data, we cannot rule out inaccuracy in our read-depth estimates of copy numbers for these clones (ie. they have 3 copies). An alternate explanation is a secondary rearrangement of an original inverted triplication resulting in a duplication (Brewer et al., 2024); however, we did not detect evidence for secondary rearrangements in the sequencing data. A third alternate explanation is that a duplication was formed by hairpin capped double-strand break repair (Narayanan et al., 2006). Notably, we found 3 additional ODIRA clones that end in native telomeres, each of which had amplified 3 copies. In these clones the other breakpoint contains the centromere, indicating the entire right arm of chromosome XI was amplified 3 times via ODIRA, each generating supernumerary chromosomes. Thus,ODIRA can result in amplifications of large genomics regions from segmental amplifications to supernumerary chromosomes.”

(b) In Panel B the violin plots appear to indicate that there are two size categories for amplicons in the ARS∆ strain. Do clones from these different sub-populations share a common CNV architecture?

Thank you for making this point. (Please note that the violin plots are now Figure 4E) We added a short discussion and Supplementary Figure 14. In line 432, we state:

“In ARS∆, we find two CNV length groups (Figure 4E) that correspond with two different CNV mechanisms (Supplementary Figure 14). 100% of smaller CNVs (6-8kb) (Supplementary Figure 14) correspond with a mechanism of NAHR between LTRs flanking the GAP1 gene (Figure 4H, ARS∆, bottom left green points). Larger CNVs (8kb-200kb) (Supplementary Figure 14) correspond with other mechanisms that tend to produce larger CNVs, including ODIRA and NAHR between one local and one distal LTR element (Figure 4H).”

(c) Panels D and E: There is great information in these two panels but I find the color keys confusing. There doesn't seem to be any reason for the strain color key in panel E. I am assuming that the key should go with Panel D. Is there some way to indicate in Panel D which events are in which CNV category? It is cumbersome to find that information from Panel E. Perhaps the color-coding from Panel E could be applied to the row labels in Panel D. Being able to link amplicon to the mechanism of CNV formation is especially important for seeing which ODIRA events contain an origin.

Thank you for this suggestions. We now indicate the mechanism of CNV formation using a consistent color coding in panels G and H (previously panels D and E).

(d) Panel E: I don't understand the two axes in Panel E. If both axes are log scales, why is the origin 0 for the X-axis and 1 for the Y-axis? And why are the focal amplicons (most of which are recombination events between the two LTRs) scattered in both X and Y coordinates? Shouldn't they form a single point? The same for the recombinants with distal LTRs. Also, orange and red (ODIRA and complex CNVs, respectively) are very hard to distinguish. All of these data need to be presented in a spreadsheet identifying each clone's strain ID, chemostat number, GAP1 and GFP copy numbers, sequence across the junction, and their coordinates. The SRA project (PRJNA1016460) for the sequence data was not found in SRA. Will this data be available to easily look at read depth across chromosome XI for all of the sequenced strains - perhaps as .bam files?

Thank you for calling these issues with data visualization to our attention. Indeed, the focal amplifications do form around a single point. We originally had jittered the data to show each individual focal amplification but agree that this is confusing. We now overlay the individual points and have altered opacity to enable visualization of individual values. The suggested table of clone data is provided in Supplementary File 2 and the SRA project is now publicly available. Moreover, we are providing all alignment (.bam) files, split, and discordant read depth profiles for each CNV strain and their corresponding ancestor aligned to our custom reference genomes in a public jbrowse server at:

https://jbrowse.bio.nyu.edu/gresham/?data=data/ee_gap1_arch_muts for WT strains, https://jbrowse.bio.nyu.edu/gresham/LTRKO_clones for LTR∆ strains, https://jbrowse.bio.nyu.edu/gresham/ARSKO_clones for ARS∆ strains, https://jbrowse.bio.nyu.edu/gresham/ALLKO_clones for ALL∆ strains.

(e) Supplementary Table 1 and Supplementary Figure S2: Please indicate which rearrangements (of the 8 reported in Figure S2A) were identified in each of the clones described in the table. If each of the 8 amplicons is identified by a letter, then this information could be added as a column in the table. I am assuming that each of the eight rearrangements was found in more than one chemostat. Showing these data is crucial for establishing the possibility that they were preexisting at the time of chemostat inoculation. The other possibility is that the clones with amplified GAP1 but a single copy of GFP could have been created by a secondary rearrangement in the outgrowth of the clones that originally had amplified both genes to the same extent. What is the structure of these amplicons? Is there a common junction between GAP1 and GFP? I couldn't find these data in the paper. A suggestion for Supplemental Figure S2A - include a zoomed-in inset for the GAP1 GFP region for each of the 8 read-depth plots. It is hard to see the exact location of GFP and GAP1 across all 8 tracks without getting out a ruler. Were these sequences aligned to your custom reference genome or the reference genome without GFP? If they were aligned to the custom reference that includes the GFP reporter, the reader could visually confirm the absence of GFP amplification.

Thank you for these suggestions. We edited Supplementary Table 1 and Supplementary Figure 1A as requested. We now provide the precise CNV breakpoints in the GFP-GAP1 region (supplemental figure 1B) displaying both genome read depth and split read depth tracks. These sequences were aligned to the custom reference containing the GFP reporter, which is now clearer in the figure and caption text in line 1226.

The clones in this figure were sampled from the five different chemostats and we have clarified this in the edited table and text at line 210. We did not detect the same CNV allele in different chemostats and therefore we do not have evidence to support GAP1 amplification without the GFP reporter pre-existing at time of inoculation. We are not able to definitively distinguish whether the amplicons were pre-existing at the time of inoculation or occurred after as we do not have barcoded lineages. We isolated clones carrying this class of amplification from the 1-GFP-copy subfraction late in the experimental evolution (generation 165-182). Given that the alleles appear to differ between populations we think the most parsimonious explanation is that these amplifications occurred after chemostat inoculation but early in the evolution experiment. We explicitly state this in the text starting in line 219.

(9) Page 8-9: I am sorry to say that I can't evaluate the "HDI of posterior distributions". It is out of my competency range. So I am not sure what this analysis is adding to the paper. The same goes for the rest of the supplementary figures.

HDI is a measure of certainty in an estimate, similar to confidence interval. We state this in the text in line 276. With the editing of the text we hope the modeling and its supplementary figures are more clear now.

(10) Page 9 top: Deletion of the ARS appears to lower the fitness of the amplified GAP1 variants. Can the authors speculate on why the ARS deletion would reduce fitness? Did they consult published replication profiles to determine the size of the origin-free gap that could result from the deletion of this mid-S phase origin? Could it explain the delay in the appearance of GAP1 amplicons in the ARS-deletion strains and be responsible for their reduced selection coefficients? Did you examine the growth properties of the starting strain or any of the amplified GAP1 derivatives? Perhaps this consideration could contribute to the discussion. Could there be a bit fuller discussion on the interaction between CNV length differences as shown in Figure 4A and differences in selection coefficient as determined by the nnSBI?

Thank you for raising this point. We have now added text to our discussion of the reduced fitness in ARS∆ in relation to DNA replication starting on line 359:

“ARS1116 is a major origin (McGuffee et al., 2013) and ODIRA CNVs found around this origin corroborate its activity. GAP1 is highly transcribed in glutamine-limited chemostats (Airoldi et al., 2016). Head-on transcription-replication collisions at this locus may be contributing to the higher CNV formation rate in wild type and LTR∆. Elimination of the local ARS could result in less transcription-replication collisions and the slower CNV formation rates estimated. Once formed they get outcompeted by faster-forming CNVs and thus in theory are less fit than CNVs in other strain backgrounds. These simulated competitions further suggest that the ARS is a more important contributor to adaptive evolution mediated by GAP1 CNVs.”

We examined replication profiles in McGuffee et al. Mol Cell. 2013 but could not determine the size of the origin-free gap. ARS1116 and its neighboring ARSs, ARS1118 downstream and ARS1115 upstream are efficient firing origins (Supplement 1 of McGuffee et al. 2013) and therefore the gap is likely to be minimal. The dynamics of the distal firing ARS elements involved in creating ODIRA CNVs might explain the reduced fitness, but further experiments would be required to address this. Regarding growth properties, the growth rate at steady-state in the chemostat is the same as the dilution rate regardless of strain background. Because we had the same dilution rate for each chemostat, the ARS∆ populations would have the same replication rate as the other three strains even if there may be replication rate differences in bulk culture growth. Finally, we found no significant interaction between CNV length and selection coefficients and we state this in line 359.

(11) Page 10: WT competition simulations: It may help to explicitly state that the competition modeling approach was experimentally validated in Avecilla 2022 as opposed to just citing the paper. I found the results much more convincing after reading Avecilla 2022, but I imagine many readers may skip that.

We added a sentence to state that the nnSBI method was experimentally validated in Avecilla et 2022 at line 249.

Reviewer #2 (Recommendations For The Authors):

(1) Figure 2: says reported CNV proportions (dashed). This may be a typo since I think the GFP reported should be solid, not dashed. Also, (C) isn't bold.

Thank you for identifying these mistakes. We have corrected the figure’s caption in line 1157.

(2) "compared to 898/345 clones" Does this refer to transposition/clone? Seems more natural to compare clones with transpositions to a total number of clones. This could be clarified.

We rephrased the sentence (lines 519-520) to clarify that in their study Hays et al. 2023 found 898 novel Ty insertions across 345 nitrogen-evolved clones. As a result of this high rate of transposition, some clones are expected to have multiple Ty insertions.

(3) The methods state that Kan replaces the Nat cassette that was used to make the deletions. It should be made more clear whether Kan is present and where Kan is with respect to GFP and GAP1.

Thank you for pointing this out. To clarify we added the following sentence to the methods starting in line 567:

“The CNV reporter is 3.1 kb and located 1117 nucleotides upstream of the GAP1 coding sequence. It consists of, in the following order, an ACT1 promoter, mCitrine (GFP) coding sequence, ADH1 terminator, and kanamycin cassette under control of a TEF promoter and terminator.”

Additionally in line 571 we clarify the drug resistance of the genomic architecture ∆ strains that are kanamycin(+) and nourseothricin(-).

Reviewer #3 (Recommendations For The Authors):

(1) The major advancement of the manuscript is stated in the title "DNA replication errors are a major source of adaptive gene amplification" First, in my humble opinion the term replication errors is not quite right; the term template switching is more accurate. In that regard, recently a paper was published just on this topic (Martin et al Plos Genetics, 2024).

We have changed the title to “Template-switching during DNA replication is a prevalent source of adaptive gene amplification”. We cite Martin et al Plos Genetics 2024 throughout the main text in lines 93, 126, 159, 502, 555.

(2) I find the statement "We find that 49% of all GAP1 CNVs are mediated by the DNA replication-based mechanism Origin Dependent Inverted Repeat Amplification (ODIRA) regardless of background strain." Somewhat misleading, there were considerable differences between the strains. If I am not mistaken the range was 20-80%.

Thank you for pointing this out. Indeed, the range was 26-80% across the four strains. We updated this sentence in the abstract at line 40, and in the main text at line 141 to clearly state the range.

(3) In their attempt to fill the gap of knowledge regarding the fitness effect of the adaptive CNV the authors use a mathematical model. As an experimental biologist, I found the description lacking. It is hard for me to evaluate the contribution of the model to understanding the results and I think the authors could improve this part.

We have edited the text regarding the modeling and associated results and hope that it is now more clear. The mathematical model describes the experiment in a simplified manner. We use it to predict the outcomes of additional experiments without additional experimental work. For example, we used it to simulate a competition between two strains, predict the total proportion of GAP1 CNVs, and predict the relative genetic diversity.

(4) Experiments the authors may want to consider to increase the novelty of their work:

a) Place the GAP1 gene right in the middle of the two most distant ARS elements and test the mechanism of CNV.

Thank you for this proposed experiment. It is beyond the scope of this paper and will be pursued in future studies.

b) The finding of de-novo Ty element insertion is interesting. What happens if the overdose strain of Jef Boeke is used (Retrotransposon overdose and genome integrity, PNAS 2009) or in contrast, a reverse transcriptase deficient strain?

We agree. Our study has revealed a critical role for novel Ty insertion in mediating CNVs. The suggested experiments as well as using strains that lack Ty sequences will be very interesting to explore in followup studies.

c) The genomic analyses were based on single colony isolates. To my understanding, the CNV events are identified at least partly by split reads. Therefore, each event may have a "signature" that is unique and can be concluded from single reads and not necessarily from the assembled genome. If true, a distinction between the scenarios could be achieved if bulk cultures are sequenced with enough depth. Thus, a truly dynamic and quantitative determination of the different events, rate of appearance, and disappearance can be made.

Thank you for this suggestion, which is a good idea but not currently feasible for several reasons. First, although split reads are a powerful way to detect CNV breakpoints, we have found that even at high coverage (21-153X, median 78.5X), in clonal samples that are rare with only 3-30 split reads (median 14) detected. These observations are from a total of 23 breakpoints across 16 sequenced clones. Thus, when sequencing heterogeneous cultures, in which different CNVs only comprise a fraction of the population, our ability to detect single CNV alleles by split reads and quantify their frequency is limited. Given our observations, with a median of 14 split reads when sequencing to 78.5X genome-wide read coverage it is possible we may be able to detect an individual CNV allele once it makes up (14/78.5) 17% of the population. However, our previous study has shown that there are tens to hundreds of unique CNV alleles initially and thus this would only be feasible at very late timepoints. Second, recurrent CNVs may occur independently at the same exact location, such as LTR NAHR. Thus, unique signatures may not be obtained even if they are independent events. Third, it would be not appropriate to pursue this analysis with our current dataset, as we lack lineage tracking barcodes to validate the results.

Version 2

Abstract

Abstract

Copy number variants (CNVs)—gains and losses of genomic sequences—are an important source of genetic variation underlying rapid adaptation and genome evolution. However, despite their central role in evolution little is known about the factors that contribute to the structure, size, formation rate, and fitness effects of adaptive CNVs. Local genomic sequences are likely to be an important determinant of these properties. Whereas it is known that point mutation rates vary with genomic location and local DNA sequence features, the role of genome architecture in the formation, selection, and the resulting evolutionary dynamics of CNVs is poorly understood. Previously, we have found that the GAP1 gene in Saccharomyces cerevisiae undergoes frequent and repeated amplification and selection under long-term experimental evolution in glutamine-limiting conditions. The GAP1 gene has a unique genomic architecture consisting of two flanking long terminal repeats (LTRs) and a proximate origin of DNA replication (autonomously replicating sequence, ARS), which are likely to promote rapid GAP1 CNV formation. To test the role of these genomic elements on CNV-mediated adaptive evolution, we performed experimental evolution in glutamine-limited chemostats using engineered strains lacking either the adjacent LTRs, ARS, or all elements. Using a CNV reporter system and neural network simulation-based inference (nnSBI) we quantified the formation rate and fitness effect of CNVs for each strain. We find that although GAP1 CNVs repeatedly form and sweep to high frequency in strains with modified genome architecture, removal of local DNA elements significantly impacts the rate and fitness effect of CNVs and the rate of adaptation. We performed genome sequence analysis to define the molecular mechanisms of CNV formation for 177 CNV lineages. We find that across all four strain backgrounds, between 26% and 80% of all GAP1 CNVs are mediated by Origin Dependent Inverted Repeat Amplification (ODIRA) which results from template switching between the leading and lagging strand during DNA synthesis. In the absence of the local ARS, a distal ARS can mediate CNV formation via ODIRA. In the absence of local LTRs, homologous recombination mechanisms still mediate gene amplification following de novo insertion of retrotransposon elements at the locus. Our study demonstrates the remarkable plasticity of the genome and reveals that template switching during DNA replication is a frequent source of adaptive CNVs.

Introduction

Defining the genetic basis and evolutionary dynamics of adaptation is a central goal in evolutionary biology. Mutations underlying adaptation or biological innovation can depend on multiple factors including genetic backgrounds, phenotypic states, and genome architecture (Blount et al., 2008, 2012). One important class of mutation mediating adaptive evolution are copy number variants (CNVs) which comprise duplications or deletions of genomic sequences that range in size from gene fragments to whole chromosomes. Quantifying the rates at which CNVs occur, the factors that influence their formation, and the fitness and functional effects of CNVs is essential for understanding their role in evolutionary processes.

CNVs play roles in rapid adaptation in multiple contexts and are an initiating event in biological innovation. For example, in laboratory evolution experiments a spontaneous tandem duplication captured a promoter for expression of a citrate transporter and resulted in Escherichia coli cells, typically unable to use citrate, to start metabolizing citrate as a carbon source (Blount et al., 2012). CNVs can be beneficial in cancer cells, promote tumorigenesis (Ben-David & Amon, 2020), enhance cancer cell adaptability (Rutledge et al., 2016), and accelerate resistance to anti-cancer therapies (Lukow et al., 2021). Over longer time scales, CNVs serve as substrate from which new genes evolve (Ohno, 1970; Taylor & Raes, 2004) as duplicated genes redundant in function can accumulate mutations and evolve to acquire new functions. For example, the globin gene family in mammals arose from rounds of gene duplication and subsequent diversification (Storz, 2016). CNVs also contribute to macro-evolutionary processes and thereby contribute to species differences, such as between humans and chimpanzees (Cheng et al., 2005) and reproductive isolation (Zuellig & Sweigart, 2018).

Mutations, including CNVs, occur in part because of errors made during DNA replication or DNA repair. Two general processes underlie CNV formation: (1) DNA recombination-based mechanisms and (2) DNA replication-based mechanisms (Brewer et al., 2011; Harel et al., 2015; Hastings, Lupski, et al., 2009; Malhotra & Sebat, 2012; Pös et al., 2021; F. Zhang, Gu, et al., 2009). Recombination-mediated mechanisms of CNV formation include non-allelic homologous recombination (NAHR) and nonhomologous end joining. NAHR occurs via recombination between homologous sequences that are not allelic. As such, NAHR occurs more frequently with repetitive sequences due to improper alignment of DNA segments and can occur either between (interchromosomal) or within (intrachromosomal) a chromosome (Harel et al., 2015). One prevalent class of repetitive sequence are retrotransposons and both full length and partial sequences, such as long terminal repeats (LTR), are substrates for homologous recombination generating gene amplifications (Avecilla et al., 2023; Dunham et al., 2002; Gresham et al., 2008; Lauer et al., 2018; Spealman et al., 2022). DNA replication-based mechanisms include fork stalling template switching (FoSTes) and microhomology mediated break-induced repair (MMBIR) (Carvalho et al., 2013; Gu et al., 2008a; Hastings, Ira, et al., 2009; Lee et al., 2007). During FoSTes and MMBIR, the DNA replication fork stalls due to a single strand nick and a replication error occurs in which the lagging strand switches to an incorrect template strand mediated by microhomology. Reinitiation of DNA synthesis at the incorrect site can form CNVs. A particular type of DNA replication-based error is Origin-Dependent Inverted Repeat Amplification (ODIRA), in which short inverted repeats near an origin of DNA replication enable template switching of the leading strand to the lagging strand. Subsequent replication generates an intermediate DNA molecule that can recombine into the original genome to form a triplication with an inverted middle copy (Brewer et al., 2011, 2015; Martin et al., 2024).

In microbes, CNVs can mediate rapid adaptation to selective conditions imposed through nutrient limitation in a chemostat. Selected CNVs often include genes encoding nutrient transporters that facilitate import of the limiting nutrient (Dunham et al., 2002; Gresham et al., 2008; Hong & Gresham, 2014; Horiuchi et al., 1963; Payen et al., 2016; Sonti & Roth, 1989), likely as a result of improved nutrient transport capacity due to increased protein production. Previous studies have found amplification of the general amino acid permease gene, GAP1, when Saccharomyces cerevisiae populations are continuously cultured in glutamine-limited chemostats (Gresham et al., 2010; Lauer et al., 2018). Amplification of GAP1 confers increased fitness in the selective environment (Avecilla et al., 2023). Sequence characterization of these CNVs revealed that a diversity of de novo CNV alleles are generated and selected including tandem duplications, complex large CNVs, aneuploidies, and translocations. However, little is known about the molecular mechanisms underlying this diversity.

Local genome sequence elements are likely to be an important determinant of CNV formation rates and mechanisms. Genomic context can influence multiple properties including mutation rate, epigenetic regulation, chromatin state, transcription levels, DNA replication, and recombination rate (Arndt et al., 2005; Chuang & Li, 2004; Lang & Murray, 2011; Lercher & Hurst, 2002; Matassi et al., 1999; Nishant et al., 2009; Wolfe et al., 1989). Prior work has shown that CNVs occur more frequently in repetitive regions in the genome (Harel et al., 2015; Pentao et al., 1992; Stankiewicz et al., 2003; Turner et al., 2008). However, little is known about the role of local genomic architecture and organization on CNV formation rates, the types of CNVs that are generated, their associated fitness effects, and ultimately the paths taken during adaptive evolution.

Here, we aimed to investigate the effect of local genome architecture elements on de novo GAP1 CNV formation and selection dynamics during adaptive evolution of Saccharomyces cerevisiae. We hypothesized that sequence elements proximate to GAP1 potentiate CNV formation. The GAP1 locus, which is located on the short arm of chromosome XI, consists of two flanking Ty1 long terminal repeats (LTRs) that share 82% sequence identity and an origin of DNA replication or autonomously replicating sequence (ARS) (Figure 1A). Both LTRs and ARS may facilitate GAP1 CNV formation due to their proximity. First, the flanking LTRs can undergo inter-chromatid NAHR to form tandem duplications of GAP1 on a linear chromosome (Lauer et al., 2018; Spealman et al., 2022). Second, intra-chromatid NAHR between the flanking LTRs can form an extrachromosomal circle containing GAP1 and an ARS able to self-propagate and integrate into the genome (Gresham et al., 2010). Finally, GAP1 triplications can form through ODIRA using short inverted repeats and the proximate ARS (Brewer et al., 2015; Lauer et al., 2018; Martin et al., 2024). These elements are thought to facilitate a high rate of GAP1 amplification, estimated to be on the order of 10-4 per haploid genome per generation (Avecilla et al., 2022). To test our hypothesis we used a CNV reporter, wherein a constitutively expressed fluorescent GFP gene is inserted adjacent to GAP1 (Lauer et al., 2018). We engineered strains that lacked either the ARS (ARSΔ), both flanking LTRs (LTRΔ), or all three elements (ALLΔ) (Figure 1A). We performed experimental evolution using wildtype (WT) and genomic architecture mutant populations in glutamine-limited chemostats for 137 generations and quantified GAP1 CNVs using flow cytometry (Figure 1). Surprisingly, we find that the proximate DNA elements are not required for GAP1 CNV formation as GAP1 CNVs were identified in all evolving populations. We used neural network simulation-based inference (nnSBI) to infer the CNV formation rate and selection coefficient (Avecilla et al., 2022). We find that although genomic architecture mutants have significantly reduced CNV formation rates relative to WT and significantly lower selection coefficients, GAP1 CNVs repeatedly form and sweep to high frequency in all strains with modified genomes. We performed genome sequence analysis to define the molecular mechanisms of CNV formation for 177 CNV lineages and found that 26-80% of GAP1 CNVs are mediated by ODIRA across all four background strains. In the absence of the local ARS, a distal ARS facilitates CNV formation through ODIRA. We also find that homologous recombination mechanisms still mediate gene amplification in the absence of LTRs in part initiated by de novo insertion of retrotransposon elements at the locus. Our study reveals the remarkable plasticity of the genome and that template switching of the leading and lagging strands during DNA replication is a common source of adaptive CNVs even in the absence of local DNA sequences.

A local DNA replication origin contributes to CNV dynamics during adaptive evolution.

(A) The Saccharomyces cerevisiae GAP1 gene is located on the short arm of chromosome XI (beige rectangle). Light blue rectangle - Ty Long terminal repeats (LTR). Purple rectangle - Autonomously replicating sequences (ARS). Orange rectangles - tRNA genes. GAP1 ORF - white rectangle. The GAP1 gene (white rectangle) is flanked by Ty1 LTRs (YKRCδ11, YKRCδ12), which are remnants of retrotransposon events, and is directly upstream of an autonomously replicating sequence (ARS1116). Variants of the GAP1 locus were engineered to remove either both LTRs, the single ARS, or all three elements. All engineered genomes contain a CNV reporter. (B) We evolved the four different strains in 5-8 replicate populations, for a total of 27 populations, in glutamine-limited chemostats and monitored the formation and selection of de novo GAP1 CNVs for 137 generations using flow cytometry. Population samples were taken every 8-10 generations and 100,000 cells were assayed using a flow cytometer. Colored lines show the median proportion of cells in a population with GAP1 amplifications across 5-8 replicate populations of the labeled strain. The shaded regions represent the median absolute deviation across the replicates. (C) We summarized CNV dynamics and found that strain has a significant effect on CNV appearance (Kruskal-Wallis, p =0.001384). There are significant differences in CNV appearance between LTRΔ (blue) and ARSΔ (red), and LTRΔ (blue) and ALLΔ (yellow) (pairwise wilcoxon test with Bonferroni correction, p=0.0059 and p=0.0124, respectively). (D) Strain has a significant effect on the per generation increase in proportion of cells with CNV (ANOVA, p = 0.00318) calculated as the slope during CNV selection phase. There is a significant difference between LTRΔ (blue) and ALLΔ (yellow) (pairwise t-test with Bonferroni correction, p = 0.0026). (E) Strain has a significant effect on time to CNV equilibrium phase (ANOVA, p = 0.00833). There is a significant difference in time to CNV equilibrium between WT and ARSΔ (pairwise t-tests with bonferroni correction, p =0.050).

Results

Accurate estimation of CNV allele frequencies remains challenging using molecular methods such as DNA sequencing and qPCR. To address this challenge we previously developed a CNV reporter comprising a constitutively expressed fluorescent gene inserted upstream of GAP1 and observed recurrent amplification and selection of GAP1 in glutamine-limiting chemostats (Lauer et al., 2018). Subsequently, we showed that a high rate of GAP1 CNV formation and strong fitness effects explain the highly reproducible evolutionary dynamics (Avecilla et al., 2022). Noncoding sequence elements proximate to GAP1, including flanking LTRs in tandem orientation and an ARS, contribute to GAP1 CNV formation (Gresham et al., 2010; Lauer et al., 2018). Many studies have shown that repetitive sequence regions and origins of replications are hotspots of CNVs (Arlt et al., 2012; Cardoso et al., 2016; Di Rienzi et al., 2009; Gresham et al., 2010; Lauer et al., 2018; Martin et al., 2024; H. Zhang et al., 2013). Thus, we hypothesized that the local genomic architecture of GAP1 facilitates its high rate of CNV formation.

To test the role of proximate genomic features we engineered strains in which each element is deleted and thus differ from the wildtype strain (WT) containing a GAP1 CNV reporter by a single modification. Specifically, we constructed ARSΔ, a strain lacking the single ARS, LTRΔ, a strain lacking the flanking LTRs, and ALLΔ, a strain lacking all three elements (Figure 1A). All strains contain the CNV reporter at the identical location as the WT strain. We confirmed scarless deletions of genetic elements using Sanger and whole-genome sequencing.

Local genomic architecture contributes to GAP1 CNV evolutionary dynamics

We founded independent populations with each of the three engineered strains lacking proximate genomic features and a WT strain. We studied GAP1 CNV dynamics in populations maintained in glutamine-limited chemostats over 137 generations (Figure 1). For each of the four strains, we propagated 5-8 clonal replicate populations, each originating from the same inoculum (founder population) derived from a single colony. Approximately every 10 generations, we measured GFP fluorescence of sampled populations using a flow cytometer and quantified the proportion of cells containing GAP1 CNVs (Methods). We observed similar CNV dynamics across independent populations within each strain (Figure S3B). Therefore, we summarized CNV dynamics for each strain using the median proportion of the population with a GAP1 CNV (Figure 1B). In every strain, GAP1 CNVs are generated and selected resulting in qualitatively similar dynamics in WT and mutant strains.

Deletion of the ARS, but not the flanking LTRs alters CNV dynamics

We quantified three phases of CNV dynamics: 1) time to CNV appearance, defined by the inflection point before the rise in CNV proportion (Figure 1C); 2) selection of CNV, corresponding to the increase in proportion of CNVs per generation during the initial expansion of CNVs (i.e., slope) (Figure 1D); and 3) equilibrium phase, corresponding to the inflection point before the plateau (Figure 1E). The time to CNV appearance (Figure 1C) and the CNV selection (Figure 1B) does not differ between WT and LTRΔ populations (pairwise Wilcoxon test, adjusted p = 1, pairwise t-test adjusted p = 1, respectively). In the WT and LTRΔ populations, GAP1 CNVs appear at generation 50 (Figure 1C) and increase in proportion at similar rates, ∼15% per generation in WT and ∼18% per generation in LTRΔ (Figure 1D). The two strains both reach their equilibrium phase at the same time, around generation 75 (pairwise t-test, adjusted p =1) (Figure 1E). The absence of a significant difference in CNV dynamics between the two strains suggests that the LTRs are not a major determinant of GAP1 CNV evolutionary dynamics.

By contrast, in ARSΔ and ALLΔ populations, we observe a delay in the time to CNV appearance. In both of these strains, CNVs are first detected at generations 65-80, whereas in WT and LTRΔ populations CNVs are first detected at generation 50 (ARSΔ vs. LTRΔ, wilcoxon pairwise test, adjusted p = 0.0059, ALLΔ vs. LTRΔ, Wilcoxon pairwise test, adjusted, p = 0.0124) (Figure 1C). Thus, the local ARS contributes to the initial GAP1 CNV dynamics. Similarly, CNV selection is significantly different between the LTRΔ (18%) and ALLΔ (13%) (pairwise t-test, adjusted p-value = 0.0026) (Figure 1D). Finally, we also observe a significant delay (ANOVA, p = 0.00833) in the generation at which the CNV proportion reaches equilibrium in ARSΔ (∼generation 112) compared to WT (pairwise t-test, adjusted p = 0.05) (Figure 1E). These observations suggest that absence of the ARS in the ARSΔ and ALLΔ strains delays the appearance of GAP1 CNVs compared with the presence of the ARS in WT or LTRΔ strains.

GAP1 amplifications can occur without CNV reporter amplification

In both WT and LTRΔ populations we observed that GAP1 CNV abundance stabilized around 75% during the equilibrium phase (Figure 1B) across each of the twelve independent populations (Supplementary S3B). Flow cytometry analysis showed that each experiment begins a population of cells with only one-copy of GFP (Figure 2A). Over generations, distinct populations appear with higher GFP fluorescence (Figure 2A). Previously, GFP fluorescence has been shown to scale with GAP1 copy number (Lauer et al., 2018). Therefore, the four distinct subpopulations observed (Figure 2A) likely represent cells harboring 1, 2, 3, and 4-copies of GAP1, respectively. This corroborates previous experimental evolution results in which de novo GAP1 CNVs are quickly formed and selected for and over selection higher copy outcompete lower copy subpopulations (Lauer et al., 2018). The raw flow cytometry plots (Supplementary Figure 2) and population GFP histograms (Supplementary Figure 3) also revealed a persistent single-copy GFP subpopulation throughout the timecourse (Figure 2A, bottom subpopulation in each panel). These data could be explained by two possible scenarios: 1) the existence of a non-GAP1 CNV subpopulation comprising beneficial variation at other loci with fitness effects equivalent to GAP1 CNVs; or 2) lineages with GAP1 CNVs without co-amplification of the CNV reporter. To resolve these two possibilities, we sequenced clones from the single-copy GFP subpopulation across of the five WT populations from different chemostats (Supplementary Table 1) and identified the presence of GAP1 amplifications without co-amplification of the CNV reporter in four out of five WT populations (Supplementary Figure 1). We found eleven distinct GAP1 CNVs that lacked amplification of the reporter gene (Supplementary Figure 1) indicating at least eleven independent CNV events occurred, either in the founder population or shortly after chemostat inoculation. By contrast, in one of the five populations, population 3, all clones from the single-copy GFP subpopulation contained one copy of GFP and one copy of GAP1 (Supplementary Table 1), suggesting these clones have a beneficial mutation elsewhere in the genome that allows their stable coexistence with the GAP1 CNV subpopulation. The GAP1 CNVs without GFP amplification were either pre-existing at the time of the inoculation or occurred shortly after inoculation. We suspect that they likely occurred after inoculation but early in the evolution, for three reasons: 1) the similar ∼75% plateau is observed in the dynamics in all independent WT and LTRΔ populations, 2) at least eleven independent CNVs of this type were detected (GAP1 amplification without co-amplification of the GFP), and 3) there are no common CNVs detected across chemostats. Our findings show that GAP1 amplification without coamplification of the CNV reporter can occur and beneficial variation other than GAP1 CNVs underlie adaptation to glutamine-limitation.

CNV reporter failure does not impact parameter inference.

(A) Flow cytometry of a representative WT population with a persistent one-copy GFP subpopulation, bottom in each panel. FSC-A is forward scatter-area which is a proxy for cell size (x-axis). GFP fluorescence was measured in arbitrary units (a.u.) (y-axis). Hierarchical gating was performed to define the one-copy GFP and two-or-more copy subpopulations (see Methods). (B) Model illustration. XA is the frequency of ancestor cells in the chemostat; XC+, XC− are the frequencies of cells with GAP1 duplications with two or one reporters, respectively, and a selection coefficient sC; XB is the frequency of cells with other beneficial mutations and a selection coefficient sB. GAP1 duplications form with a rate δC, other beneficial mutations occur with rate δB. At generation 0, only genotypes C− and A are present, with frequencies of XC− = φ and XA = 1 − φ. (C) Examples of total CNV proportions (dashed) and reported CNV proportions (solid) for two parameter combinations, both with s = 0. 15, φ = 10−4.

Incorporating unreported early-occurring CNVs in an evolutionary model

To quantify the evolutionary parameters underlying empirically measured CNV dynamics (Figure 1) we built a mathematical evolutionary model, which describes the experiment in a simplified manner. Because measuring CNV rates and selection coefficients is difficult and laborious to perform in the lab, we use neural network simulation-based inference (nnSBI) to estimate these parameters (Avecilla et al., 2022; Cranmer et al., 2020; Gonçalves et al., 2020). Additionally, we use the model to predict hypothetical outcomes of additional experiments without requiring additional experimentation. We have previously used nnSBI to infer GAP1 CNV formation rates and selection coefficients in glutamine-limited selection and experimentally validated these inferences using barcode tracking and pairwise competition assays (Avecilla et al., 2022). Previously, our evolutionary model assumed the GAP1 CNV reporter allowed us to detect all GAP1 CNVs. However, our new flow cytometry and sequencing results indicate the existence of a small subpopulation of unreported GAP1 CNVs (Figure 2B, C- cells) present either at the beginning or early in the experiments. Therefore, we expanded the evolutionary model to include φ, the proportion of cells with GAP1 CNVs without co-amplification of the reporter, at the commencement of the experiment (i.e., generation 0). The remaining model parameters are δC, the rate at which GAP1 duplications form; δB, the rate other beneficial mutations occur; sC, the selection coefficient of GAP1 CNVs; and sB, the fitness effect of other beneficial mutations (Figure 2B). We find that this expanded evolutionary model can accurately describe the observed dynamics (Supplementary Figure 5), which are clearly affected by the value of φ. When the total CNV proportion is very different from the reported proportion, e.g., when φ≫δC > δB, a reduced CNV formation rate results in a greater discrepancy between reported and total CNV proportions (Figure 2C).

Decreased CNV formation rates in modified genomes suggests adjacent genomic elements contribute to GAP1 CNV formation

We used nnSBI to infer CNV formation rates and selection coefficients from the evolutionary dynamics observed in glutamine-limited chemostats (Figure 1B). Previously, nnSBI estimations have been experimentally validated demonstrating its accuracy and reliability (Avecilla et al., 2022). First, we trained a neural density estimator using evolutionary simulations (Methods). This neural density estimator then allows us to infer posterior distributions and estimate the model parameters (i.e. the GAP1 CNV formation rate, δC; the GAP1 CNV selection coefficient, sC) from a single population CNV dynamics. We also inferred a collective posterior distribution from a set of replicate populations of the same strain. This collective posterior distribution consolidates estimations of the formation rate and selection coefficient from multiple replicate populations of one strain into one single estimate per strain in order to compare between the four strains, rather than between all 27 populations. We evaluated the confidence of our inference approach on synthetic simulations by computing its coverage, i.e., the probability that the true parameter falls within the 95% highest density interval (HDI) of the posterior distribution, a measure of certainty in an estimate similar to confidence intervals (Kruschke, 2021)(Supplementary Table 2). We find that the posterior distributions are narrow as the 95% HDI are less than an order of magnitude for both sC and δC. Thus, we did not apply post-training adjustments to the neural density estimator, such as calibration (Cook et al., 2006) or ensembles (Caspi et al., 2023; Hermans et al., 2022) when estimating δC and sC from experimental GAP1 CNV dynamics.

We find that the individual maximum a posteriori (MAP) estimates vary across strains and replicates (Supplementary Figure 4). Overall, the CNV selection coefficient, sC, ranges from 0.1 to 0.22 (with one exception of 0.3), whereas the CNV formation rate, δC, ranges from 10−6 to 10−4 (with one exception of 10−3 and two of 10−7); and the proportion of early-occurring GAP1 CNVs without amplification of the reporter (φ) ranges from 10−6 to 10−2 (with two exceptions of 10−8). We found that MAP estimates of replicate populations of the same strain cluster together, with some outliers (Supplementary Figure 4). We performed posterior predictive checks, drawing parameter values from the posterior distributions and simulating the CNV dynamics (Supplementary Figure 10), which agree with the observed data (Supplementary Figure 5). For each strain, we use all individual posterior distributions to infer the collective posterior distribution, which is a posterior distribution conditioned on all observations, P(θ|X1, …, Xn) (Methods). The collective posterior allows us to estimate whether there is a difference in CNV formation rate and fitness effect across the four strains.

Collective posterior HDIs are very narrow (Figure 3A), and samples are highly correlated, as expected for joint estimation of selection coefficients and beneficial mutation rates (Gitschlag et al., 2023). The collective MAP estimates of the CNV selection coefficient are similar for the WT and LTRΔ (0.182). For ARSΔ and ALLΔ, the selection coefficient is estimated to be lower, with values of 0.146 and 0.126, respectively. However, all four selection coefficients are still large, consistent with these populations containing GAP1 CNVs that are highly beneficial under glutamine-limitation. The collective MAP estimate for the CNV formation rate in WT is 4. 5⋅10−5. By contrast, the CNV formation rate is markedly lower in all mutant strains ranging from 1⋅10−5 for LTRΔ and ALLΔ to 2. 4⋅10−6 in ARSΔ. These results support our hypothesis that proximate sequence features facilitate GAP1 CNV formation.

Inference of CNV formation rate and selection coefficient from experimental evolutionary data.

(A) Collective MAP estimate (black markers) and 50% HDR (colored areas) of GAP1 CNV formation rate, δC, and selection coefficient, sC. Marginal posterior distributions are shown on the top and right axes. (B) Collective posterior prediction of Shannon diversity of CNV lineages (e−Σi[pilog(pi)], Jost, 2006). Line and shaded area show mean and 50% HDI. (C) CNV reported frequency (XC+) prediction using collective MAP (solid line) compared to empirical observations (dotted lines).

The collective MAP predictions reproduce the experimental observations. Other than the very final time point for the WT population, all collective MAP predictions lay within the interquartile ranges (Supplementary Figure 5). The observed GAP1 CNV proportion stabilizes at different levels in the different experiments (Figure 1B). This can be explained by pre-existing or early-occurring unreported CNVs with proportions estimated to be between φ=4⋅10−6 to 1. 6⋅10−4 by the collective MAPs (Supplementary Figure 6). Indeed, our model predicts that the total (reported and unreported) final CNV proportion is nearly one in all cases (Supplementary Figure 9).

We sought to understand the consequences of differences in CNV formation rate and CNV fitness effects between the four strains on evolutionary dynamics. We used a modified version of the evolutionary model with the estimated parameters to simulate an evolutionary competition between WT and the three architecture mutant strains over 116 generations, a point at which CNVs have reached high proportions in the experiment. To “win” these competitions, the competitor strains need to adapt to glutamine-limitation by producing CNVs. The results of the simulated competitions predict that the WT outcompetes the other strains in all cases as its predicted final proportion almost always exceeds its initial proportion of 0.5 (Supplementary Figure 11 and supplementary information). The average predicted proportion of WT cells when competing with LTRΔ is 0.717. By contrast, ARSΔ and ALLΔ are predicted to be almost eliminated by generation 116, as the average predicted WT proportion is 0.998 and 0.999, respectively. These simulated competitions further suggest that the ARS is a more important contributor to adaptive evolution mediated by GAP1 CNVs.

Next, we estimated de novo CNV diversity in each strain. Previous work showed a diversity of CNV alleles formed under glutamine-limited selection including tandem duplications, segmental amplification, translocations, and whole chromosome amplification (Lauer et al., 2018), and that lineage richness decreases rapidly over the course of evolution due in part to competition and clonal interference (Lauer et al., 2018; Levy et al., 2015; Nguyen Ba et al., 2019). Our model does not include competition, clonal interference, or recurrent CNV formation. Therefore, diversity calculations are likely overestimations. Nonetheless, a comparison of diversity between strains is informative of whether proximate genome elements affect CNV allele diversity. Therefore, for each strain, we used its collective MAP to simulate a posterior prediction for the genotype frequencies (Figure 3C), which we then used to predict the posterior Shannon diversity (Jost, 2006). In all populations, we predict the set of CNV alleles to be highly diverse: the final predicted Shannon diversity ranges from 1. 6⋅104 in ARSΔ to 3. 2⋅105 in WT (Figure 3B). Our model predicts that the diversity increases rapidly during the selection phase and stabilizes in the equilibrium phase. This is because CNV alleles that form towards the end of the experiment would have a low frequency with a minor effect on diversity. We observe the greatest diversity in WT populations with lower diversity in the three genomic architecture mutants. Moreover, diversity saturates faster in WT populations. This suggests that the WT strain is able to form more unique CNVs allele types earlier compared to the other three strains (Figure 3B). Shannon diversity is lower in LTRΔ and further lower in ALLΔ and ARSΔ (Figure 3B) reflecting the rank order of CNV formation rates (Figure 3A).

Inference of CNV mechanisms in genome architecture mutants

Contrary to our expectations, removal of proximate genomic elements from the GAP1 locus does not inhibit the formation of GAP1 CNVs. We sought to determine the molecular basis by which GAP1 CNVs form in the absence of these local elements. Therefore, we isolated 177 GAP1 CNV-containing clones across each population containing the four different strains at generations 79 and 125 and performed Illumina whole-genome sequencing. Using a combination of read depth, split read, and discordant read analysis, we defined the extent of the amplified region, the precise CNV breakpoints, and GAP1 copy number. On the basis of these features, we inferred the CNV-forming mechanisms for each GAP1 CNV (Methods). Among the 177 analyzed GAP1 CNVs, we observed tandem amplifications, tandem triplications with an inverted middle copy, intra- and inter-chromosomal translocations, aneuploidy, and complex CNVs. GAP1 copy numbers range from two to six in any given clone. Each of the four strains is able to produce a diversity of CNV alleles ranging from small (tens of kilobases) to large (∼hundreds of kilobases) segmental amplifications (Figure 4). We quantified the CNV length per strain (Figure 4E) and found no significant interaction between CNV length and generation from which the clone was isolated (ANOVA, p=0.33) and therefore considered all 177 clones in subsequent comparisons (Supplementary Figure 12). We found no significant effect of the inferred sC (ANOVA, p= 0.673) or δC (ANOVA, p= 0.277) on CNV length. We defined six major CNV-forming mechanisms across the four strains: ODIRA, LTR NAHR, NAHR, transposon-mediated, complex CNVs, and whole chromosome duplication (aneuploidy) and assigned each CNV allele to one mechanism using diagnostic features of each CNV (Figure 4A-D and Methods).

Local and distal elements contribute to generation of GAP1 CNV alleles.

(A) Schematic of Saccharomyces cerevisiae GAP1 locus on Chromosome XI: 513332-518060 with LTR, ARS elements and tRNA genes labeled. Long terminal repeat non-allelic homologous recombination (LTR NAHR) is defined on the basis of both CNV breakpoints occuring at LTR sites as revealed by read depth plots (left, pink and blue vertical lines) and increased read depth relative to the genome-wide read depth (left). In some cases we detect a hybrid sequence between two LTR sequences, a result of recombination between the two LTRs (right). LTR NAHR typically forms tandem duplications.

(B) ODIRA is a DNA replication-error based CNV mechanism generated by template switching of the leading and lagging strand template at short inverted repeats. In this clone example, the relative read depth estimate of 2.67 copies of GAP1 is rounded to 3 copies (left) and has apparent breakpoints in the CAF4 and AIM29 genes. We classify a clone as being formed by ODIRA if it has a de novo inverted sequence in at least one breakpoint. In the clone example, the short inverted repeat pairs are CAAT, ATTG (ChrXI:508938, ChrXI: 508986) in CAF4 and TAAAA, AAAAT (ChrXI:582561, ChrXI582610) in AIM29. The contig sequence at the breakpoint (rectangle) is aligned to the reference within the CAF4 coding sequence fragment. The 5’ and 3’ ends of the contig are labeled and a dashed line indicates contiguity (no gaps). The contig spans the CAF4 breakpoint junction and contains a de novo inversion, i.e.) two fragments of the CAF4 gene are in opposite orientations with the mediating short inverted repeats shown in blue and underlined. The contig was generated using CVish (Methods) and supported by split reads at the breakpoint junction (not shown).. The contig containing a de novo inversion across the AIM29 breakpoint is not shown. ODIRA typically forms tandem triplications with an inverted middle copy and contains an ARS (bottom).

(C) Non-allelic homologous recombination (NAHR) is defined by having at least one CNV breakpoint not at the proximate LTR sites, ie. other homologous sequences in the genome. In the example clone, we detect a hybrid sequence between the two homologous sequences in BET3 (ChrXI) and SPT4 (ChrVI). Because these two sequences are on different chromosomes we infer that an interchromosomal translocation occurred. The other breakpoint is unresolved. A read depth plot supports the amplified segment containing the GAP1 gene. NAHR is able to produce supernumerary chromosomes as is the case in Clone 2968 (Figure 4G)

(D) Transposon-mediated mechanism is defined by an inference of at least one intermediate novel LTR retrotransposon insertion followed by LTR NAHR. In the ALLΔ strains which have the flanking LTRs deleted, we find novel LTR retrotransposon insertions near previously deleted LTR sites. The newly deposited LTR sequence (downstream of GAP1) recombines with another LTR sequence (upstream of GAP1), either pre-existing or introduced by a second de novo retrotransposition, to form a resulting CNV (focal amplification or segmental amplification). Read depth estimation (not shown) supports the CNV breakpoints at pre-existing or newly deposited LTRs.

(E) Violin plot of CNV length in each genome-sequenced clone, n = 177. Strain has a significant effect on CNV length, Kruskal-Wallis test, p = 1.008 x 10-5. Pairwise wilcoxon rank sum test with bonferroni correction show significant CNV length differences between WT and LTRΔ (p = 0.00490), WT and ALLΔ (p = 0.01230), LTRΔ and ARSΔ (p=0.00056), ARSΔ and ALLΔ (p=0.002).

(F) Barplot of inferred CNV mechanisms, described in A-D, for each CNV clone isolated from glutamine-limited evolving populations. Complex CNV is defined by a clone having more than two breakpoints in chromosome XI, indicative of having more than one amplification event. Inference came from a combination of read depth, split read, and discordant read analysis and the output of CVish (see Methods). Strain is significantly associated with CNV Mechanism, Fisher’s Exact Test, p = 5.0 x 10-4. There is a significant increase in ODIRA prevalence between WT and LTRΔ, chi-sq, p = 0.02469. There is a significant decrease in ODIRA prevalence from WT to ARSΔ and ALLΔ, chi-sq, p = 0.002861 and 0.002196, respectively. There is a significant decrease of LTR NAHR from WT to LTRΔ, chi-sq, p = 0.03083.

(G) Top: Schematic of S. cerevisiae chromosome XI, with LTR, ARS elements, tRNA genes annotated. LTR-blue, ARS-purple, tRNA-orange, GAP1 ORF-white rectangle. Using a combination of read depth, split read, and discordant read analysis, we defined the extent of the amplified region, the precise CNV breakpoints, and GAP1 copy number. GAP1 copy numbers were estimated using read depth relative to the average read depth of chromosome XI. Bottom: Dumbbell plots represent the amplified region (>1 copy) relative to the WT reference genome. The ends of the dumbbells mark the approximate CNV breakpoints shown relative to the start codon of the GAP1 ORF (vertical dotted line). Select clones were chosen as representative of the observed diversity of amplifications.

(H) Scatterplots of CNV length for all genome-sequenced clones, n = 177. We defined the upstream and downstream breakpoints as kilobases away from the start codon of the GAP1 ORF (vertical dashed line in (G) dumbbell plot and this scatterplot). CNV mechanisms are defined in Figure 4A-D and Methods. Select clones from (G) dumbbell plots are annotated. Note in the focal amplications resulting for LTR NAHR in WT clones and ARSΔ clones, respectively. In ARSΔ, note NAHR between one local and one distal LTR ∼60 kb upstream. Note in ALLΔ focal amplications mediated by two newly deposited LTR sequences from two transposon insertions. Note in ALLΔ amplifications formed between one newly inserted LTR and one distal pre-existing LTR sequence, 30kb or 60 kb upstream.

ODIRA is a predominant mechanism of CNV formation

We inferred GAP1 CNVs formed through ODIRA in all four genotypes at high frequencies: 22 out of 37 WT clones (59%), 42 out of 52 LTRΔ clones (81%), 11 out of 42 ARSΔ clones (26%), and 12 out of 46 ALLΔ clones (26%). Considering the set of all CNVs in all strains, ODIRA is the most common CNV mechanism comprising almost half of all CNVs (87/177, 49%). The second most common mechanism occurs about half as often—NAHR between flanking LTRs (38/177, 21%), which generates tandem amplifications. In the WT background, ODIRA (22/37) and NAHR between LTRs (11/37) account for 89% of GAP1 CNVs.

In LTRΔ populations, GAP1 CNVs form via ODIRA, chromosome missegregation, and NAHR using other sites. As expected, in LTRΔ clones we did not detect NAHR between LTRs in 52 clones and no focal amplifications were detected (Figure 4H). In LTRΔ populations CNVs are formed predominantly by ODIRA (42/52, 81%) (Supplementary Table 3), a significant increase relative to WT clones (chi-sq, p = 0.02469) (Figure 4F). By contrast, aneuploidy (5/52), complex CNV (3/52), and NAHR (2/52) account for less than 10% of GAP1 CNVs in LTRΔ. Consequently, we observe an increase in average GAP1 CNV length in LTRΔ relative to WT (Figure 4E) as there is an increased prevalence of segmental amplifications and aneuploidy (Figure 4H).

Overall, aneuploidy was observed infrequently. Whole amplification of chromosome XI was detected in six out of 177 clones (3.4%) (Figure 4F) and detected in only two strains: WT and LTRΔ (Figure 4G). We also detected supernumerary chromosomes in five out of 177 clones (2.8%), which formed through both NAHR and ODIRA (Figure 4G).

ODIRA generates CNVs using distal ARS

Whereas removal of proximate LTRs prevents the formation of small tandem duplication CNVs through LTR NAHR, removal of the local ARS does not prevent the formation of GAP1 CNVs through ODIRA (Figure 4F). In the absence of the proximate ARS, distal ones are used to form ODIRA as all amplified regions of ODIRA clones contain a distal ARS (Figure 4G), with one exception (Methods, Supplementary Figure 13). We observe a significant increase of LTR NAHR in the ARSΔ clones (27/52, 52%) relative to WT clones (11/37, 39%) (Figure 4F, chi-sq, p = 0.03083). In ARSΔ clones, we find two CNV length groups (Figure 4E) that correspond with two different CNV mechanisms (Supplementary Figure 14). All smaller CNVs (6-8kb) (Supplementary Figure 14) correspond with a mechanism of NAHR between LTRs flanking the GAP1 gene (Figure 4H, ARSΔ, bottom left green points). Larger CNVs (8kb-200kb) (Supplementary Figure 14) correspond with other mechanisms that tend to produce larger CNVs, including ODIRA and NAHR between one local and one distal LTR element (Figure 4H).

Surprisingly, we found CNVs with breakpoints consistent with ODIRA that contained only 2 copies of the amplified region, whereas ODIRA typically generates a triplication. In the absence of additional data, we cannot rule out inaccuracy in our read-depth estimates of copy numbers for these clones (ie. they have 3 copies). An alternate explanation is a secondary rearrangement of an original inverted triplication resulting in a duplication (Brewer et al., 2024); however, we did not detect evidence for secondary rearrangements in the sequencing data. A third alternate explanation is that a duplication was formed by hairpin capped double-strand break repair (Narayanan et al., 2006). Notably, we found 3 additional ODIRA clones that end in native telomeres, each of which had amplified 3 copies. In these clones the other breakpoint contains the centromere, indicating the entire right arm of chromosome XI was amplified 3 times via ODIRA, each generating supernumerary chromosomes. Thus,ODIRA can result in amplifications of large genomics regions from segmental amplifications to supernumerary chromosomes.

Novel retrotransposition events potentiate GAP1 CNVs

CNVs in the ALLΔ clones form by two major mechanisms: 1) ODIRA using distal ARS sites to form large amplifications and 2) LTR NAHR following novel Ty LTR retrotransposon insertions to form focal amplifications (transposon-mediated, Figure 4). These two classes are evident in the broad CNV lengths detected (Figure 4E). ALLΔ clones tend to have more larger amplifications formed by ODIRA than ODIRA-generated amplifications in WT and LTRΔ (Figure 4H) because they encompass distal ARS and inverted repeats (Figure 4G). Surprisingly, we detected novel LTR retrotransposon events that generated new LTRs that subsequently formed GAP1 CNVs through NAHR with a pre-existing LTR in the genome or an LTR from a second novel retrotransposition (Figure 4H). This explains the small focal amplifications detected in ALLΔ clones that are in some cases smaller than that of WT (Figure 4E). Regions upstream of tRNA genes are known to be hotspots for Ty retrotransposons (Ji et al., 1993; Mularoni et al., 2012). We find the novel retrotransposons insert near one or both of the previously deleted LTR sites (Supplementary File 1), which flank GAP1 and are downstream of tRNA genes (Figure 1A). We only detected novel retrotranspositions in ALLΔ populations. In total we detected 15 unique Ty retrotransposon insertion sites of which eight were upstream of the deleted LTR, YKRCδ11, and four were downstream of deleted LTR, YKRCδ12 (Supplementary File 1). The remaining two insertions were distal to the GAP1 gene: one on the short arm and the second on the long arm of chromosome XI. Every novel insertion was upstream of an tRNA gene, consistent with the biased preference of Ty LTR insertions (Ji et al., 1993; Mularoni et al., 2012). Recombination between a new and preexisting LTR produces large amplifications whereas recombination between two newly inserted Ty1 flanking the GAP1 gene forms focal amplifications of the GAP1 gene (Figure 4H).

Discussion

In this study we sought to understand the molecular basis of repeated de novo amplifications and selection of the general amino acid permease gene, GAP1, in S. cerevisiae evolving under glutamine-limited selection. We hypothesized that a high formation rate of GAP1 CNVs is due to the unique genomic architecture at the locus, which comprises two flanking long terminal repeats and a DNA replication origin. We used genetic engineering, experimental evolution, and neural network simulation-based inference to quantify de novo CNV dynamics and estimate the CNV formation rate and selection coefficient in engineered mutants lacking the proximate genome elements. We find that removal of these elements has a significant impact on de novo CNV dynamics, CNV formation rate, and selection coefficients. However, CNVs are formed and selected in the absence of these elements highlighting the plasticity of the genome and diversity of mechanisms that generate CNVs during adaptive evolution.

Despite their proximity to GAP1 and previous studies demonstrating the prevalence of NAHR between repetitive sequences forming CNVs (Dunham et al., 2002; Gresham et al., 2010; Todd et al., 2019; H. Zhang et al., 2013), we found that flanking LTRs are not an essential driver of CNV formation. The de novo CNV dynamics of WT and LTRΔ populations are similar and we find that although the CNV formation rate is reduced, the effect is small. By contrast, a significantly decreased CNV formation rate and delayed CNV appearance time was observed in the absence of the ARS in ARSΔ and ALLΔ populations, which suggests that the local ARS is an important determinant of GAP1 CNV-mediated adaptive dynamics. Furthermore, the significant delay in the time at which the CNV frequency reaches equilibrium in the ARSΔ compared to WT (Figure 1E) can be explained by both the estimated lower CNV formation rate and lower selection coefficient (Figure 3A). ODIRA was identified as the predominant CNV mechanism in sequence-characterized clones revealing that DNA replication errors, specifically template switching of the leading and lagging strands, are a common source of CNV formation during adaptive evolution.

Using nnSBI we inferred lower rates of CNV formation in all strains with modified genomes that may be informative of the rate at which specific mechanisms occur. The lower CNV formation rate in the LTRΔ strain could be a closer approximation of ODIRA formation rates at this locus as ODIRA CNVs are the predominant CNV mechanism in the LTRΔ strain (Figure 4F). Furthermore, the low formation rates in the LTRΔ relative to WT might suggest that the presence of the flanking long terminal repeats may increase the rate of ODIRA formation through an otherwise unknown combinatorial effect of DNA replication across these flanking LTRs and template switching at the GAP1 locus. ARSΔ has the lowest CNV formation rate and it could be an approximation of the rates of NAHR between flanking LTRs and ODIRA at distal origins. We find that the ALLΔ has a higher CNV formation rate than the ARSΔ. One explanation for this is that the deletion of the flanking LTRs in ALLΔ gives opportunity for novel transposon insertions and subsequent CNV formation through LTR NAHR. Indeed, we find an enrichment of novel transposon-insertions in the ALLΔ (Figure 4F) and subsequent CNV formation through recombination of the Ty1-associated repeats (Figure 4H, ALLΔ). The sequential events of transposon insertion followed by LTR NAHR must occur at a high rate to explain the increased CNV rate in ALLΔ relative to ARSΔ. While remarkable, increased transposon activity is associated with nutrient stress (Curcio & Garfinkel, 1999; Lesage & Todeschini, 2005; Todeschini et al., 2005) and therefore this is a plausible explanation for the CNV rate estimated in ALLΔ. Additionally, ARSΔ clones rely more on LTR NAHR to form CNVs (Figure 4F). The prevalence of ODIRA in ARSΔ and ALLΔ are similar. LTR NAHR usually occurs after double strand breaks at the long terminal repeats to give rise to CNVs (Argueso et al., 2008). Because we use haploid cells, such double strand break and homology-mediated repair would have to occur during S-phase after DNA replication with a sister chromatid repair template to form tandem duplications. Therefore the dependency on LTR NAHR to form CNVs and the spatial (breaks at LTR sequences) and temporal (S-phase) constraints could explain the lower formation rate in ARSΔ.

The genomic elements have clear effects on the evolutionary dynamics using simulated competitive fitness experiments. The similar selection coefficients in WT and LTRΔ suggest that CNV clones formed in these background strains are similar. Indeed, the predominant CNV mechanism in both is ODIRA followed by LTR NAHR (Figure 4F). Whereas LTR NAHR is abolished in the LTRΔ, it seems that CNVs formed by ODIRA allow adaptation to glutamine-limitation similar to WT. The lower selection coefficients in ARSΔ and ALLΔ suggest that GAP1 CNVs formed in these strains have some cost. In a competition, they would get outcompeted by CNV alleles in the WT and LTRΔ background (Supplementary Figure 11). Additionally, the local ARS, ARS1116, is a major origin (McGuffee et al., 2013) and ODIRA CNVs found around this origin corroborate its activity. The simulated competitions (Supplementary Figure 11) further suggest that the ARS is a more important contributor to adaptive evolution mediated by GAP1 CNVs.

The prevalence of ODIRA generated CNVs is a consequence of multiple DNA replication origins and pervasive inverted repeat sequences throughout the chromosome (Figure 4). In particular, breakpoint analysis of LTRΔ CNV clones show that ODIRA produces a continuum of CNV sizes along the short arm of chromosome XI. Downstream breakpoints of ODIRA-generated CNVs range from nearby the GAP1 gene (∼3 kilobases) to the right telomere of chromosome XI (153 kilobases) (Figure 4H). The S. cerevisiae genome contains a high frequency of inverted repeats ranging from 3bp to 14bp throughout the genome (Martin et al., 2024), but longer repeats are more likely to be used in ODIRA (Martin et al., 2024). The ubiquity of inverted repeats is in stark contrast to the relative paucity of LTR sequences, which are dispersed throughout the genome. Thus, ODIRA supplies a diverse and high number of gene amplifications for selection to act on, setting the stage for genome evolution and adaptation. It appears that complex CNVs may include secondary rearrangements after an initial ODIRA event as most complex CNVs in this study include signatures of ODIRA events (Supplementary Table 2). Further work needs to be done to resolve the CNV structure.

Consistent with previous reports of increased Ty insertions in S. cerevisiae under stress conditions (Morillon et al., 2000, 2002), we observed novel retrotransposon insertions in populations evolved in glutamine-limited chemostats. Transposon insertions can be harmful and lead to loss-of-function mutations but are also a means of generating beneficial alleles including CNVs (Blanc & Adams, 2003; Dunham et al., 2002; Gresham et al., 2008; Wilke & Adams, 1992). We only detected novel Ty insertions in the ALLΔ strain. This is likely because regions upstream of tRNA genes are predisposed to transposition. Our detection of novel retrotransposon insertions is consistent with a previous experimental evolution study that suggested that Ty insertions were rare under constant nitrogen-limitation and substantially more common under fluctuating nitrogen limitation, in which cells experience total nitrogen starvation periodically (Hays et al., 2023). In that study, 898 novel Ty insertions were found across 345 clones (Hays et al., 2023) corresponding to an average of 2.6 insertions per genome This high insertion frequency is consistent with detecting novel insertions on either side of the GAP1 gene that subsequently mediate a focal amplification via LTR NAHR. Importantly, the role of Ty differs in the two studies, as in our case beneficial CNV formed after novel retrotransposition through recombination of newly introduced repeat sequences, whereas Hays et al. (2023) found Ty-associated null alleles that are beneficial in nitrogen-limited conditions. Together, these results reveal the different means by which retrotransposition can facilitate adaptive evolution..

Aneuploidy was not a major source of adaptation in our experiments as it was infrequently detected (n = 6/177). This contrasts with studies suggesting aneuploidy is a rapid and transient route to adaptation over short evolutionary time scales (Chen, Bradford, et al., 2012; Chen et al., 2015; Chen, Rubinstein, et al., 2012; Pavelka et al., 2010; Selmecki et al., 2006; Selmecki et al., 2015; Yona et al., 2012). However, aneuploidy incurs a fitness cost (Robinson et al., 2023; Tsai et al., 2019; Yang et al., 2021) and therefore can be outcompeted by slow-forming but less costly beneficial mutations in large populations (Kohanovski et al., 2024). Our observed higher frequencies of focal and segmental amplifications may be because they are less costly than whole-chromosome amplifications.

A variety of DNA replication errors generate CNVs. Replication slippage at palindromic DNA and DNA repeats can cause fork stalling and downstream CNV formation (Lee et al., 2007; F. Zhang, Khajavi, et al., 2009). DNA repeats can form secondary structures like R loops, cruciforms, non-B DNA structures, and hairpins which stimulate CNV formation (Gu et al., 2008). Untimely replication, faulty fork progression, S-phase checkpoint dysfunction, defective nucleosome assembly, and DNA repeat sites including LTRs are sources of replication-associated genome instability (Aguilera & García-Muse, 2013).

Additional processes may also play a role. The GAP1 gene is highly transcribed under glutamine-limitation (Airoldi et al., 2016) and transcription-replication collisions may fuel ODIRA CNV formation at this locus (Lauer & Gresham, 2019; Wilson et al., 2015). CNV formation can also be stimulated by transcription-associated replication stress and histone acetylation (Hull et al., 2017; Salim et al., 2021; Whale et al., 2022) and replication fork stalling at tRNA genes (Osmundson et al., 2017; Yeung & Smith, 2020). Testing the role of transcription in promoting the formation of adaptive CNVs warrants further investigation.

Recent work has proposed that ODIRA CNVs are a major mechanism of CNVs in human genomes (Brewer et al., 2011, 2015; Martin et al., 2024). Studies of human and yeast genomes have typically considered homologous recombination as the predominant mechanism of CNV formation (Lupski & Stankiewicz, 2005). CNV hotspots identified in the human (Chance et al., 1994; Lupski, 1998; Lupski & Stankiewicz, 2005; Pentao et al., 1992) and yeast genomes are indeed mediated by NAHR of long repeat sequences (Green et al., 2010; Gresham et al., 2010). However, a focus on recombination-based mechanisms as a means of generating copy number variation may be the result of ascertainment bias or the comparative ease of studying the effect of long repeat sequences over short palindromic ones. Our study demonstrates that experimental evolution in yeast is a useful approach to elucidating the molecular mechanisms by which DNA replication errors generate CNVs.

Methods

Strains and Media

Each of the three architecture mutants were constructed independently starting with the GAP1 CNV reporter strain (DGY1657). The CNV reporter is 3.1 kb and located 1117 nucleotides upstream of the GAP1 coding sequence. It consists of, in the following order, an ACT1 promoter, mCitrine (GFP) coding sequence, ADH1 terminator, and kanamycin cassette under control of a TEF promoter and terminator. To construct each deletion strain, we performed two rounds of transformations both using PCR amplified donor templates designed for homology-directed repair. The first transformation used a repair template containing a nourseothricin resistance cassette to replace the pre-existing kanamycin resistance cassette and GAP1 gene. The repair template was designed to also delete the elements of interest (ie. ARS, both flanking LTRs, or both LTRs and ARS). The second transformation replaced the nourseothricin cassette with a kanamycin resistance cassette and GAP1 gene thus yielding a genomic architecture Δ strain that is kanamycin(+) and nourseothricin(-). We confirmed scarless deletions with sanger sequencing and whole-genome-sequencing. Final identifiers are DGY1657 for the WT strain, DGY2076 for the LTRΔ strain, DGY2150 for the ARSΔ strain, and DGY2071 for the ALLΔ strain.

The zero-, one-, and two-copy GFP controls, DGY1, DGY500, and DGY1315, respectively, are described in Lauer et al. 2018 and Spealman et al. 2023 (Lauer et al., 2018; Spealman et al., 2023). Briefly, GFP under the ACT1 promoter was inserted at neutral loci that do not undergo amplification in glutamine-limited continuous culture. 400μM glutamine-limited media is described in Lauer et al. 2018.

Long-Term Experimental Evolution

We performed experimental evolution of 30 S. cerevisiae populations in miniature chemostats (ministats) for ∼137 generations under nitrogen limitation with 400μM glutamine as in Lauer et al. (2018). Of the 30 populations, there were three controls: one control population with no fluorescent reporter (DGY1), one with one GFP fluorescent reporter (DGY500), one with two GFP fluorescent reporters (DGY1315). The remaining 27 populations have the GAP1 CNV GFP reporter. Of these, five populations are WT (DGY1657), seven are LTRΔ (DGY2076), seven are ARSΔ (DGY2150), and eight are ALLΔ (DGY2071). We inoculated each ministat containing 20ml of glutamine-limited media with 0.5 ml culture from its corresponding genotype founder population. The founder population was founded by a single colony grown overnight in glutamine-limited media at 30°C. Replicate populations of the same strain were inoculated from the same founder population derived from a single colony. Strains were randomized among the 30-plex ministat setup to account for the possibility of systematic position effects. After inoculation, populations were incubated in a growth chamber at 30°C for 24 hours with the media inflow pump off. After 24 hours, the populations had reached early stationary phase and we turned on the media inflow pump and waited 4 hours for the populations to reach steady-state equilibrium, at which the population size was ∼108 cells. This was generation zero. Ministats were incubated in a growth chamber at 30°C with a dilution rate of 0.12 culture volumes/hr. Since the ministats had a 20ml culture volume, the population doubling time was 5.8 hours. Approximately every 10 generations, we froze 2-ml samples of each population in 15% glycerol stored at −80°C. Approximately every 30 generations, we pelleted cells from 1-ml samples of each population and froze them at −80°C for genomic DNA extraction.

Flow Cytometry analysis to study GAP1 CNV dynamics

To track GAP1 CNV dynamics, we sampled 1-ml from each population approximately every 10 generations. We sonicated cell populations for 1 minute to remove any cell clumping and immediately analyzed samples on the Cytek Aurora flow cytometer. We sampled 100,000 cells per population and recorded forward scatter, side scatter, and GFP fluorescent signals for every cell. We performed hierarchical gating to define cells, single cells, unstained (zero-copy-GFP control) cells, cells with one copy of GFP (GAP1), and two or more copies of GFP (GAP1) (Spealman et al., 2023). First we gated for cells (filtered out any debris, bacteria) by graphing forward scatter area (FSC-A) against side scatter area (SSC-A). Second, we gated for single cells by graphing forward scatter area against forward scatter height and drawing along the resulting diagonal. Finally, we drew non-overlapping gates to define three subpopulations: zero copy, one copy, and two or more copies of GFP by graphing B2 channel area (B2-A), which detects GFP (excitation = 516 λ, emission = 529 λ), against forward scatter area (FSC-A). We note that the one copy and two copy events overlap some, which is a limitation in this experiment (Spealman et al., 2023).

We found that two architecture mutants, DGY2150 and DGY2071, had strain-specific GFP fluorescence even though they only harbored one copy of GFP. DGY2150 and DGY2071 had slightly higher fluorescence than the one copy GFP control strain, DGY500, but less than that of the two copy GFP control strain, DGY1315. The third architecture mutant, DGY2076, had the same GFP fluorescence as the one-copy GFP control strain (DGY500). We ruled out that they were spontaneous diploids by looking at forward scatter signals. The forward scatter signal was not different from that of the one copy control (a haploid) and was not as high as a diploid. Therefore due to strain-specific fluorescence, we decided to perform strain-based gating, ie. one set of gates for the WT strain, a second set of gates for the LTRΔ strain, and so on. Since the controls are also a strain of their own, they were not used to set universal gates for one-copy or two-copy. Thus, for each strain, we chose the basis of our one-copy gate as the timepoint per strain in aggregate with the lowest median cell-sized normalized fluorescence. The two-or-more-copy (CNV) gate was drawn directly above and non-overlapping with the one-copy gate.

Quantification of dynamics

To obtain the proportion of CNVs for each population at each timepoint, we applied gates that correspond to zero-,one-, and two-or-more copy subpopulations. Using such proportion per population per timepoint, we summarize population CNV dynamics as follows (Lauer et al., 2018; Spealman et al., 2023). We calculate the generation of CNV appearance for each of the evolved populations. We defined CNV appearance as the generation where the proportion of CNV-containing cells first surpasses a threshold of 10% for three consecutive generations. Next, modified from Lang et al. (2011) (Lang et al., 2011) and Lauer et al. (2018) (Lauer et al., 2018), we calculate the percent increase in CNVs per generation for each evolved population. We compute the natural log of the proportion of the population with CNVs divided by the proportion of the population without CNVs for each timepoint. These proportions were obtained previously by gating. We plot these values across time and perform linear regression during the initial increase of CNVs. The slope of the linear regression is the percent increase in CNVs per generation. Finally, we calculate the time to CNV equilibrium, as defined by the generation at which a linear regression results in a slope < 0.005 after the selection phase.

Neural network simulation-based inference of evolutionary parameters

Evolutionary model

We developed a Wright-Fisher model that describes the evolutionary dynamics, similar to our previous study (Avecilla et al., 2022). In that study we have shown that a Wright-Fisher model is suitable for describing evolutionary dynamics in a chemostat. Wright Fisher is a discrete-time evolutionary model with a constant population size and non-overlapping generations. Every generation has three stages: selection, in which the proportion of genotypes with beneficial alleles increases; mutation, in which genotypes can gain a single beneficial mutation or CNV; and drift, in which the population of the next generation is generated by sampling from a multinomial distribution. Our model follows the change in proportion of four genotypes (Figure 2B): A, the ancestor genotype; B, a cell with a non-CNV beneficial mutation; C+, a genotype with two copies of GAP1 and two copies of the CNV reporter; and C− a genotype with two copies of GAP1 but only a single copy of the CNV reporter. CNV and non-CNV alleles are formed at a rate of δC and δB and have a selection coefficient of sC and sB, respectively. The proportion of genotype i is Xi. Unlike XB and XC+, which may increase due to both mutation and selection, we assume that C− is not generated after generation 0 (as experimental results suggest that the reporter is working properly). Hence, the proportion of the C− genotype only increases idue to selection, with sC as its selection coefficient. We assume C− has an initial proportion φ. Model equations and further details are in the supplementary information.

Simulation-based inference

We use a neural network simulation-based inference method, Neural Posterior Estimation or NPE (Papamakarios, 2019) to estimate the joint posterior distribution of three model parameters, sC, δC and φ, while the other parameters, sB and δB are fixed to a specific value (Table 1). Inferring all five model parameters resulted in similar prediction accuracy and sC and δC estimates.

We applied NPE, implemented in the Python package sbi (Tejero-Cantero et al., 2020), using a masked autoregressive flow (Papamakarios et al., 2018) as the neural density estimator: an artificial neural network that “learns” an amortized posterior of model parameters from a set of synthetic simulations. Posterior amortization allows us to infer the posterior distribution P(θ|Χ) for a new observation Χ without the need to re-run the entire inference pipeline, i.e., generating new simulations and re-training the network (as is the case in sampling-based methods such as Markov chain Monte Carlo or MCMC).

We generated 100,000 synthetic observations simulated from our evolutionary model using parameters drawn from the prior distribution (Table 1). The neural density estimator was trained using early stopping with a convergence threshold of 100 epochs without decreases in minimal validation loss (the default in sbi is 20). Using 100 epochs as a threshold resulted in improved predictions. We validated that this improvement in prediction accuracy is not a result of over-fitting (Supplementary Figure 8).

We validated the trained neural density estimator by measuring the coverage property: the probability that parameters fall within the inferred posterior marginal 95% HDI. Then, we used the distribution of (Supplementary Figure 7) and posterior predictive checks (Supplementary Figure 5) as quantitative and qualitative measures of prediction accuracy, respectively.

Collective posterior distribution

NPE estimates a single posterior distribution per observation, i.e., P(θ|X). Given n observations X1, …Xn generated from the same model distribution P(θ)P(X|θ), where each observation is a time-series of GAP1 CNV proportion, NPE infers n individual posterior distributions, each conditioned on a single observation, P(θ|Xi). We infer the collective posterior distribution based on n individual posteriors, that is, a posterior distribution conditioned on all observations, This can be computed using the individual posteriors P(θ|Xi) and the prior P(θ) (see supplementary information for derivation).

However, as P(θ|Xi) could be infinitesimally small, a single observation could potentially reject a parameter value that is likely according to other observations. We want the collective posterior to be robust to such non-representative observations. Therefore, we define Pɛ(θ|Xi) = max(ɛ, P(θ|Xi)) and use this quantity instead of P(θ|Xi) in eq. 1. For a correct choice of ɛ, the collective posterior mode should reflect a value with high posterior density for multiple observations, rather than a value that no individual posterior completely rejects. We set ɛ = e−150 based on a visual grid-search. To find the normalizing factor (denominator in eq. 1), the integral is approximated by a dense Riemann sum (300 points). Maximizing the distribution, i.e., finding the collective MAP, is implemented using scipy’s minimize method with the Nelder-Mead algorithm.

Genetic diversity

Using our evolutionary model with the inferred parameters, we can estimate the diversity of CNV alleles in the experiments. For each strain, we used samples from its collective posterior to simulate a posterior prediction for the CNV allele frequencies (Figure 3C), which we then used to compute the posterior Shannon diversity (Jost, 2006), as detailed in the supplementary information.

Whole genome sequencing of isolated clones

Clones were isolated from archived populations and verified to harbor a GAP1 CNV by measuring GFP fluorescence signal consistent with two or more copies. Populations of each strain from generation 79 were streaked out from the −80°C archive on YPD and incubated at 30°C for 2 days. Plates containing single colonies were viewed under a blue light to view GFP fluorescent colonies by eye. Relative to the fluorescence of the 2 copy control strain, we picked single colonies that fluoresced as bright or brighter, reasoning that these colonies would likely contain GAP1 CNVs. Single colonies were used to inoculate cultures in glutamine-limited media and incubated at 30°C for 18 hours. The cultures were analyzed on the Cytek Aurora to verify they indeed harbored two or more copies of GAP1 based on GFP fluorescence signal. For Illumina whole genome sequencing, genomic DNA was isolated using Hoffman-Winston method. Libraries were prepared using a Nextera kit and Illumina adapters. Libraries were sequenced on Illumina NextSeq 500 platform PE150 (2 x 150 300 Cycle v2.5) or Illumina NovaSeq 6000 SP PE150 (2x 150 300 Cycle v1.5). We also used custom Nextera Index Primers reported in S1 Table Baym et al. 2015 (Baym et al., 2015).

Breakpoint analysis and CNV mechanism inference in sequenced clones

Reference genomes

We created a custom reference genome for each of the genomic architecture mutants. The custom reference genome containing the GAP1 CNV reporter in Lauer et al. 2018 (NCBI assembly R64) was modified to delete the flanking LTRs, single ARS, or all three elements.

Copy number estimation by read depth

The estimation of GAP1 copy number from read depth used is described in Lauer et al. 2018, except we searched for ≥ 1000 base pairs of contiguous sequence. CNV boundaries were refined by visual inspection.

Structural variation calling and breakpoint analysis

Whole genome sequences of clones were run through CVish version 1.0, a structural variant caller (Spealman, 2019). Structural variant calling was also done on each of the ancestor genomes: WT, ARSΔ, LTRΔ, and ALLΔ. Output .bam files containing split reads and discordant reads of evolved clones and their corresponding ancestor were visualized on Jbrowse2 or IGV to confirm locations of de novo CNV breakpoints and orientation of sequences at the novel breakpoint junctions. Novel contigs relative to the reference genome were outputted in addition to the supporting split reads that generated the contig. Blastn was used to verify orientation of contigs, namely inverted sequences used to define ODIRA (see Definitions of Inferred CNV Mechanisms). .bam files for each analyzed evolved clone and ancestors are available for view (See Data Availability).

Definitions of Inferred CNV mechanisms

We used the following liberal classifications for each CNV category. We called a clone ODIRA if we found inverted sequences in at least one breakpoint (Figure 4A). We define LTR NAHR as having both breakpoints at LTR sites (Figure 4A), evidence of recombination between the homologous LTR sequences. This mechanism typically forms tandem amplifications. In some cases, we find the hybrid sequence between two LTRs, but this is hard to recover in short-read-sequencing. We define NAHR as having breakpoints at homologous sequences, with at least one breakpoint not at an LTR sequence (Figure 4A). We define transposon-mediated as a clone having a breakpoint at a novel LTR retrotransposon site and the other breakpoint at a different LTR site (Figure 4A). Such characteristics support that the newly deposited LTR sequence recombined with another LTR sequence (either pre-existing or introduced by a second de novo retrotransposition) to form CNVs. Rarely, we are able to recover the hybrid sequence between LTR sequences even with high sequencing coverage 80-100X). We define complex CNV as having more than two breakpoints on chromosome XI and a read depth profile that suggests more than one amplification event occurred (i.e. multi-step profile). For the complex CNV clones, we were not able to resolve the CNV mechanisms due to the limitations of short-read sequencing, though most have at least one ODIRA breakpoint.